---
title: "Geometric Signs on Paleolithic Mobile Objects as Cultural Group Markers"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
zotero: "Signbase-study"
link-citations: true
---

# Introduction

A distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the grouping patterns of early modern humans [@McDonald_Harper_2016; @Pereira_Manen_2023; @Fuentes_Lucas_Robert_2019]. During this period, we also see a rise in organized settlements, more complex social organizations, and more advanced tool-making across Europe [@Boyd_Silk_2020]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian period (ca. 43,000 - 30,000 years BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by certain lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], many of which are decorated with geometric signs [@dutkiewicz2020].

In this paper, we analyze the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe during the Aurignacian period. The basis of this analysis is the idea that variations in signs can serve as an indicator of group membership and cultural similarity. This approach is supported by ethnographic and historical evidence [@texier2010howiesons; @Hodder1979; @Hodder1977; @Vanhaeren_2005] as well as similar archaeological studies of ornamental and parietal art [@Pereira_Manen_2023; @Sauvet_2018; @Rigaud_dErrico_Vanhaeren_2015; @baker2024evidence; @Lycett_2019; @Vanhaeren_d; @McDonald_Harper_2016]. In this study, we address the following questions: Are cultural groups evident in the distribution of portable geometric signs, and if so, what can they tell us about the cultural diversity and movement of Upper Paleolithic humans?

# Background

The purpose of this study is to contribute to our understanding of the patterns of cultural and genetic diversity among the early modern humans of the Upper Paleolithic. Previous works have performed cluster analysis methods upon variations in artefacts to represent and understand patterns of groups and cultural diversity, and have primarily relied upon personal ornamentation and parietal art, with few taking mobile objects into account. For example, multivariate analyses of Gravettian beads suggests an east to west cline of nine cultural groups [@baker2024evidence]. The data set consisted of 134 discrete beads types from both burial and occupation sites, as distinguished by shape and material. Seriation analysis, PCoA, and network analysis were used to identify the groups. The geographically discrete clusters identified encompassed the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and Western Mediterranean regions. The isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, is tested using a Mantel test, determining that geographic distance cannot solely account for the bead distribution.

Similar analyses (seriation, correspondence and GIS) were also performed upon personal ornament types from the Aurignacian period as a proxy for linguistic diversity. The data set consisted of 157 distinct ornament types, and the analyses identified fourteen geographically cohesive clusters sharing similar ornament type associations [@Vanhaeren_d]. The clusters identified are geographically discrete and sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. Availability of raw materials was considered ~in regards to~ the potential explanation that observed patterns are determined by ~such~. ~It was concluded that, due to observed patterns in the data that countered that explanation~, [BM: They found that] the differences in personal ornamentation could not be explained by material availability. 

Following the established precedent of using ornamentation and signs to identify cultural groups, we look to see if designs on mobile objects can be applicable here, and what it can add to the research being done. The additional dimension of mobility in this study allows us to potentially identify more geographically dispersed cultural groups, which might help visualize the migration patterns of early modern humans. Previous studies have primarily excluded mobile objects, and tend to find more geographically compact cultural groups as a result.

Our use of geometric signs in this study is supported by @Sauvet_2018, in which used quadrilateral geometric parietal signs are used to identify Upper Paleolithic cultural groups in the Cantabrian region of Spain. The authors identified and classified the signs based on variations in form, then performed a factorial correspondence analysis and ascending hierarchical clustering upon the data in order to identify two clusters, which they argued represents distinct and separate groups of humans. In their study, @Sauvet_2018 made a case for the relevance of signs as identity markers, pointing to both the relevant qualities of signs, which are similar to those explained above, and the case studies throughout history where they served a similar purpose.

In this paper, we test the hypothesis that cultural groups are discernible in a big data set of geometric signs on mobile objects. Firstly, we identify clusters using seriation analysis, network analysis, and PCoA. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use mantel tests to evaluate the null hypothesis of isolation-by-distance. 

# Methods and Materials

```{r}
## clean the data
library(tidyverse)
library(magrittr)
library(rcarbon)
library(readxl)
signbase_full <- read_csv("data/signBase_Version1.0.csv")
nsites <- signbase_full %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts <- nrow(signbase_full)
ncountries <- signbase_full %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes <- ncol(signbase_full)
signbase_full_clean <- signbase_full %>% 
  filter(site_name != "Willendorf",
         site_name != "Riparo di Fontana Nuova",
         site_name != "Muralovka",
         site_name != "Shanidar Cave",
         site_name != "Hayonim Cave",
         site_name != "El Salitre",
         site_name != "Grotte De La Princesse Pauline",
         site_name != "Šandalja II") %>% 
  dplyr::select(-other, -rectangle)
nsites_clean <- signbase_full_clean%>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts_clean <- nrow(signbase_full_clean)
ncountries_clean <- signbase_full_clean %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes_clean <- nsigntypes - 1
signbase_years <- signbase_full_clean %>% 
  drop_na(date_bp_max_min) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
  separate(date_bp_max_min,
           sep = " - ",
           c("date_bp_max", "date_bp_min"),
           remove = FALSE) %>% 
  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
                              str_extract(date_bp_max, ".*?(?=\\/)"),
                              date_bp_max)) %>% 
  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
                              str_extract(date_bp_min, ".*?(?=\\/)"),
                              date_bp_min)) %>% 
  separate(date_bp_max,
           sep = "±",
           c("date_bp_max_age",
             "date_bp_max_error")) %>% 
  separate(date_bp_min,
           sep = "±",
           c("date_bp_min_age",
             "date_bp_min_error")) %>% 
  drop_na(date_bp_max_age,
          date_bp_max_error) %>% 
  mutate(date_bp_max_age = parse_number(date_bp_max_age),
         date_bp_max_error = parse_number(date_bp_max_error))
signbase_years_cal <- 
  rcarbon::calibrate(signbase_years$date_bp_max_age,
            signbase_years$date_bp_max_error,
            verbose = FALSE) %>% 
  summary() %>% 
  data_frame()
pretty_print_dates <- 
  function(x) formatC(x, 
                      decimal.mark=".", 
                      big.mark=",", 
                      digits = 0, 
                      format = "f")
date_min <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  min()
date_max <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  max()
date_range <- date_max - date_min
date_min_pretty <- pretty_print_dates(date_min)
date_max_pretty <- pretty_print_dates(date_max)
date_range_pretty <- pretty_print_dates(date_range)
```

```{r}
groups <- list("1" = c("Abri Pataud", "El Rascaño", 
                         "Fumane", "La Viña",
                         "Lhotka", "Pod Hradem",
                         "Riparo Bombrini", "Sirgenstein Cave"),
                "2" = c("Kvasice", "Labeko Koba", 
                        "Les Rois", "Slatinice",
                        "Göpfelsteinhöhle"),
                "3" = c("Aurignac", "d'Engihoul",
                         "Istállóskő Cave", "Hohlenstein-Stadel",
                         "Gargas", "Grotte de la Verpillière I", 
                         "Mladeč", "Brassempouy",
                         "Gatzarria"),
                "4" = c("Wildscheuer", "Breitenbach",
                        "Peskő Cave", "Vindija Cave",
                        "El Castillo"),
                "5" = c("Trou Magrite",
                        "Grotte de Goyet", "Bockstein-Törle"),
                "6" = c("Trou al'Wesse", "Maisières-Canal",
                        "Les Cottés", "Le Terme Pialat"),
                "7" = c("Vogelherd","Hohle Fels",
                        "Geissenklösterle"),
               "8" = c("Tuto de Camalhot","Abri Lartet/Gorge d'Enfer",
                        "Castanet", "Le Trou Du Renard", 
                        "Grottes de Fonds-de-Forêt" ),
               "9" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
                        "Grotte du Renne", "Solutré", 
                        "Žlutava", "Spy"),
                "10" = c("Cellier", "Blanchard", 
                         "La Ferrassie", "Hornos de la Pena",
                         "Otaslavice"),
                "11" = c("La Souquette", "Shelter Birów IV",
                         "Abri de Laussel", "Abri du Poisson"))
group_df <- 
  enframe(groups, 
          name = "group", 
          value = "site_name") %>%
  unnest(cols = site_name)

lat_long_df <- signbase_full_clean %>% 
  dplyr::select(site_name, longitude, latitude) %>% 
  distinct(site_name, .keep_all = TRUE)

signbase_unique_groups <- signbase_full_clean %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(group_df) %>% 
  left_join(lat_long_df)

artifact_data <- signbase_unique_groups %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)
```

```{r}
# mantel abundance
library(vegan)
library(ade4)

set.seed(200)
euc_data <- artifact_data %>% 
  rownames_to_column("site_name") %>% 
  group_by(site_name) %>% 
  mutate(sign_count = sum(line:star)) %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(sign_count)
  
euc_dist <- dist(x = euc_data, method = "euclidean")
jac <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)
euc_mantel <- vegan::mantel(jac, euc_dist, permutations = 1000)
p_value_abundance_correlation <- round(euc_mantel$signif, 3)
r_value_abundance_correlation <- round(euc_mantel$statistic, 3)
```

## Data

```{r fig.height= 10, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Map of Palaeolithic sites containing signs, sign abundance is represented by point size, data from Dutkiewicz et al. (2020)"
#| 
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)

library(ggpubr)

abundance_data <- signbase_full_clean %>% 
  pivot_longer(cols= line:star,
               values_to = "sign_total") %>% 
  filter(sign_total != 0) %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum))
  
signbase_sf <- 
  st_as_sf(signbase_unique_groups, 
           coords = c("longitude", "latitude"),
           remove = FALSE,
           crs = 4326)
world <- ne_countries(scale = "medium", returnclass = "sf")
Europe <- world[which(world$continent == "Europe"),]

artifact_map <- ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2) +
  theme_minimal() +
  labs(size = "Sign Count")

date_graph <- ggplot(signbase_years_cal) +
  aes(x = MedianBP) +
  geom_bar(width = 150) +
  labs(x = "Median calibrated BP")

abundance_graph <- ggplot(abundance_data) +
  aes(x = sign_total, y = reorder(site_name, sign_total)) +
  geom_col() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = "Site name",
       y = "Sign count") +
  theme(axis.text.y = element_text(size = 4.25))

library(cowplot)
plot_grid(date_graph, abundance_graph, artifact_map,
          labels = "AUTO", align = "v",
          nrow = 3, scale = c(0.75, 0.75, 1.25))

```

Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020]. The creators of the database identified around 30 distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic. Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.

The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries. The database records `r nsigntypes` geometric sign type found on these artifacts. We excluded signs if they fell under the "other" category, as the wide range of possibilities that could qualify as "other" means that it would be difficult to establish a definitive connection. We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites). After those sites were excluded, all signs under the "rectangle" category were also removed as a side effect of that exclusion. Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.

These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types, as seen in @fig-site-map. All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd. We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results, and found that it was not statistically significant [@Lycett_2019] (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`)

The artifacts in question are determined to have been made in the Paleolithic era, specifically in the Aurignacien technocomplex. The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. As seen in @fig-site-map, the dates have a few outliers, but tend to cluster between 35000 - 42500 BP.

## Testing for Cultural Groupings

Seriation analysis is the arrangement of data into a linear order to reveal patterns, and is a common technique in analyzing sets of archaeological data [@Hahsler_Hornik_Buchta]. Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns by respectively.

Network analysis represents the relationship between nodes [@Mills_2017]. In this network analysis, each node is a site, and each edge is a measure of sign similarity, with the direct value represented by edge weight. The results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.

Principal coordinate decomposition analysis (PCoA) analyzes the eigenvalue decomposition of the Jaccard matrix and arranges the sites by dissimilarity [@Legendre_Legendre_2012]. This provides a visual of sign similarity as visualized by spacing and arrangement. The PCoA in this paper is calibrated to be plotted on a scatterplot, as opposed to a biplot.

## Testing Strength of Cultural Groupings

The modularity test quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. The PerMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, and a P-score.

PERMDISP is a multivariate extension of Levene's test [@Anderson_2006, @Anderson_Ellingsen_McArdle_2006], for the analysis of multivariate homogeneity of group dispersion. In this study, we perform the PERMDISP test upon the Jaccard matrix as a measure of compositional dissimilarity. We did so using the "betadisper" function from the RStudio package 'vegan', which mimics the PERMDISP2 procedure [@Vegan2001]. In this code, the non-euclidean distance matrix is reduced to principal coordinates to embed them within a Euclidean space. The results are used to produce a visual of the level of overlap between the groups in terms of sign content.

## Testing Isolation-by-Distance Hypothesis

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography. Conversely, if geographic distance cannot account for the distribution of the sign types, we would have to assume that other factors are at play, namely, movement of groups or the creation of social networks.

Mantel tests are used here to determine whether there is a correlation between sign distribution and geography. The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix. The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], producing a p-score as a result.

# Results

## Testing for Cultural Groupings

```{r}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"
## seriation
library(ggpubr)
artifact_matrix <- as.matrix(artifact_data %>% 
 mutate(across(everything(), ~replace(., . > 1, 1))))
rownames(artifact_matrix) <- rownames(artifact_data)

# Load required libraries
library(ggplot2)

# Define the mplot function
mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
  # Create the data frame
  d <- data.frame(
    x = rep(1:ncol(x), each = nrow(x)),
    y = rep(nrow(x):1, ncol(x)),
    z = as.integer(x)
  )
  
  # Add country colors to the data frame if provided
  if (!is.null(country_colors)) {
    d$country_color <- rep(country_colors, ncol(x))
  }
  
  # Use custom row/column labels if provided, otherwise use defaults
  if (is.null(col_labels)) col_labels <- colnames(x)
  if (is.null(row_labels)) row_labels <- rev(rownames(x))
  
  # Plot
  p <- ggplot(d, aes(x, y)) +
    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
    scale_fill_identity() +  # Use the fill values directly
    coord_equal(expand = FALSE) +
    scale_x_continuous(
      breaks = 1:ncol(x),
      labels = col_labels,
      position = "top"
    ) +
    scale_y_continuous(
      breaks = 1:nrow(x),
      labels = row_labels
    ) +
    theme(
      axis.text = element_text(size = rel(0.5)),
      axis.ticks = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 0.25)
    ) +
    labs(x = "Sign Type", y = "Site Name", fill = "Count", title = title) +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
  
  p
}

# Define the increase_focus function
increase_focus <- function(x, country = NULL) {
  # Reorder rows by the mean position of 1s in each row
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  row_order <- order(mcp, na.last = TRUE)
  x <- x[row_order, ]
  
  # Reorder the country vector if provided
  if (!is.null(country)) {
    country <- country[row_order]
  }
  
  # Reorder columns by the mean position of 1s in each column
  mrp <- apply(
    x, 
    MARGIN = 2, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  x <- x[, order(mrp, na.last = TRUE)]
  
  # Return the reordered matrix and country vector
  list(matrix = x, country = country)
}

# Define the concentrate function
concentrate <- function(x, country = NULL, max_iter = 100) {
  old <- x
  not_identical <- TRUE
  iter <- 0
  
  while (not_identical && iter < max_iter) {
    result <- increase_focus(old, country)
    new <- result$matrix
    country <- result$country
    not_identical <- !identical(old, new)
    old <- new
    iter <- iter + 1
  }
  
  if (iter == max_iter) {
    warning("Reached maximum iterations without convergence.")
  }
  
  list(matrix = new, country = country)
}


# get country names, for example
x_df <- 
  artifact_data %>% 
  rownames_to_column("site_name") %>% 
  left_join(signbase_full_clean %>% 
              distinct(site_name, .keep_all = TRUE) %>% 
              select(site_name,
                     country),
            by = "site_name")

countries <- x_df$country

# Generate a named vector of colors, one for each unique country
unique_countries <- unique(countries)
color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")   # Generate distinct colors
row_colors <- setNames(color_palette, unique_countries) 
countries_colors <- row_colors[countries]

# Concentrate the matrix and reorder the country vector
result <- concentrate(artifact_matrix, country = countries)
concentrated_matrix <- result$matrix
reordered_countries <- result$country

# Map countries to colors
country_colors <- row_colors[reordered_countries]

# Plot the concentrated matrix with country colors
mplot(
  concentrated_matrix,
  country_colors = country_colors
)
```

```{r}
#| label: fig-neighbor-net
#| fig-cap: "Neighbor-Net analysis of the data"
library(phangorn)
library(ggraph)
library(statnet)
jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
dm <- as.matrix(jac)
disim <- 1 - dm
disim[disim<0.5] <- 0
disim_net <- network(disim,
                    directed=F,
                    ignore.eval=F,
                    names.eval='weight')
disim_net %v% 'vertex.names' <- 
  row.names(artifact_data)

set.seed(500)

ggraph(disim_net, layout="fr") +
  geom_edge_link(aes(width = weight, 
                     alpha=weight), 
                 edge_colour = "black", 
                 show.legend=T) +
  scale_edge_width(range=c(0.5,1.5)) +
  scale_edge_colour_gradient(low = "#CCCCCC",
                             high = "#000000") +
  geom_node_point(alpha=1, size=5) +
  geom_node_text(aes(label=row.names(artifact_data)), 
                 size=6, 
                 family= "Arial", 
                 repel=T) +
  theme(text= element_text(size = 15), 
        panel.background = element_rect(fill = 'white'))
```

```{r}
#| label: fig-pcoa
#| fig-cap: "PCoA analysis of the data"
library(ecodist)
library(ape)
pcoa <- pcoa(jac)
pcoa_df <- data.frame(pcoa1 = pcoa$vectors[,1], 
                      pcoa2 = pcoa$vectors[,2])
pcoa_df <- pcoa_df %>% 
  mutate(site_name = rownames(artifact_data))
library(ggrepel)
ggplot(pcoa_df) +
  aes(x=pcoa1, 
      y=pcoa2) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2",) +
  theme(title = element_text(size = 10)) +
  geom_text_repel(aes(label = site_name), 
                  size = 2,
                  max.overlaps = 20)
```

Combining the results of the above tests allowed us to find 11 groups of sites, as shown in @fig-group-table. The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.

## Groups

```{r}
#| label: fig-group-table
#| fig-cap: "Table of the sites by group"
library(knitr)
group_1 <- c("Abri Pataud, El Rascaño, Fumane, La Viña, Lhotka, Pod Hradem, Riparo Bombrini, Sirgenstein Cave, Šandalja II")
group_2 <- c("Kvasice, Labeko Koba, Les Rois, Slatinice, Göpfelsteinhöhle")
group_3 <- c("Aurignac, d'Engihoul, Istállóskő Cave, Hohlenstein-Stadel, Gargas, Grotte de la Verpillière I, Mladeč, Brassempouy, Gatzarria")
group_4 <- c("Wildscheuer, Breitenbach, Peskő Cave, Vindija Cave, El Castillo")
group_5 <- c("Tuto de Camalhot, Abri Lartet/Gorge d'Enfer, Castanet, Le Trou Du Renard, Grottes de Fonds-de-Forêt")
group_6 <- c("Nová Dědina, Menton/Grottes du Grimaldi, Grotte du Renne, Solutré, Žlutava, Spy")
group_7 <- c("Trou Magrite, Grotte De La Princesse Pauline, Grotte de Goyet, Bockstein-Törle")
group_8 <- c("Trou al'Wesse, Maisières-Canal, Les Cottés, Le Terme Pialat")
group_9 <- c("Vogelherd, Hohle Fels, Geissenklösterle")
group_10 <- c("Cellier, Blanchard, La Ferrassie, Hornos de la Pena, Otaslavice")
group_11 <- c("La Souquette, Shelter Birów IV, Abri de Laussel, Abri du Poisson")
group_table <- data_frame("Group" = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11"), "Site Names" = c(group_1, group_2, group_3, group_4, group_5, group_6, group_7, group_8, group_9, group_10, group_11))
kable(group_table)

```

```{r fig.height= 5, fig.width=10}
#| label: fig-group-map
#| fig-cap: "Map of sites faceted and colored by group number"
  
signbase_sf_group <- 
  st_as_sf(signbase_unique_groups, 
           coords = c("longitude", "latitude"),
           remove = FALSE,
           crs = 4326)
ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf_group,
          aes(colour = as.factor(group))) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf_group,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 100,
                  size = 2) +
  theme_minimal() +
  facet_wrap(~group) +
  labs(color = "Group Number")

```

```{r fig.height= 10, fig.width=10}
#| label: fig-group-sign
#| fig-cap: "Bar plot of signs by group"
graph_data <- artifact_data %>% 
  mutate(across(everything(), ~replace(., . > 1, 1))) %>% 
 # concentrate() %>% 
  rownames_to_column("site_name") %>% 
  left_join(group_df)

sign_type_data <- graph_data %>% 
  pivot_longer(cols= notch:hashtag,
               names_to = "sign_type",
               values_to = "count") %>% 
  dplyr::filter(count != 0) %>% 
  mutate(group = parse_number(group))

x_order <- graph_data %>% 
  dplyr::select(-site_name,
         -group) %>% 
  colnames() 

ggplot(sign_type_data) +
  aes(x = sign_type) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = x_order) +
  scale_x_discrete(limits = c(x_order)) +
  scale_y_continuous(labels = NULL) +
  facet_wrap(~group, nrow = 11)

```

## Testing Strength of Cultural Groupings

```{r}
##Modularity test
library(igraph)
set.seed(500)
modul_matrix <-
  graph_from_adjacency_matrix(1 - dm, mode = "undirected")
mem <- as_membership(parse_number(signbase_unique_groups$group))
as_group <- modularity(modul_matrix, membership = mem)
```

```{r}
##permanova test
perman <- adonis2(jac~as.factor(signbase_unique_groups$group), 
                  method = "jaccard",
                  sqrt.dist = TRUE)
perman <- perman %>% 
  data_frame() %>% 
  slice_head()
perman_r <- perman$R2
perman_p <- perman$`Pr(>F)`

```

The modularity test, when run on the above groupings, resulted in a score of `r as_group`. The perMANOVA test produced an R-squared value of `r perman_r` and a P value of `r perman_p`, which tells us that the groupings are statisically significant, and that 52% of the variability in the sign distribution can be attributed to group membership. The multivariate homogeneity test of group dispersions produced @fig-perm-disp, showing that the overlap between groups is fairly minimal.

```{r fig.height= 8, fig.width=5}
#| label: fig-perm-disp
#| fig-cap: "Multivariate homogeneity of group dispersions"

set.seed(500)
dispersion <- betadisper(jac, group = signbase_unique_groups$group)
plot(dispersion, hull = FALSE, ellipse=TRUE)

```

## Testing Isolation-by-Distance Hypothesis

```{r}
## Mantel test for geography vs signs
set.seed(200)
site_distances <- dist(cbind(signbase_unique_groups$longitude, signbase_unique_groups$latitude))
mantel_distance <- vegan::mantel(jac, site_distances, permutations = 1000)
p_value_distance_correlation <- round(mantel_distance$signif, 3)
r_value_distance_correlation <- round(mantel_distance$statistic, 3)
```

```{r}
#| label: fig-distance-mantel
#| fig-cap: "Mantel correlogram showing the results of a mantel test run between a geographic distance matrix and a dissimilarity matrix for site sign types"
distance_corr <- mantel.correlog(jac, site_distances, n.class=50, break.pts=NULL, cutoff=FALSE, r.type="pearson", nperm=999, mult="holm", progressive=TRUE)
plot(distance_corr)
```

In terms of geographic correlation, there is no significant statistical correlation between the geographic distance matrix and the sign type dissimilarity matrix (P = `r p_value_distance_correlation`, Mantel R = `r r_value_distance_correlation`). Again, the results are displayed in a mantel correlogram (@fig-distance-mantel), which demonstrates that similarity and geographic distance do not experience a clearly correlated relationship.

# Discussion

In this paper, we set out to ask if identifiable and relevant patterns existed in the data, and if so, what could they tell us about the patterns of diversity and movement among Upper Paleolithic humans?

Through our various analyses, we saw that the isolation-by-distance framework could not account for the distribution of sign types. Specifically, our Mantel test (@fig-distance-mantel) found no statistically significant relationship between geographic distance and sign variation. Therefore, we must look to other factors to explain the distribution. We also saw distinct groupings in the data, as demonstrated in @fig-seriation, @fig-neighbor-net, and @fig-pcoa. We divided the sites into approximately ten groups based on sign types with a modularity score of `r as_group`. @fig-group-map shows groups with sites both clustered geographically speaking, and more distant from each other, potentially showing us cultural groups and patterns of movement.

One explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility. @Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Similarly, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity. One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.

Further research will need to be done on this topic in order to understand these patterns. In particular, the ecological and biogeographical conditions at the sites will need to be analyzed for potential connections, as mobility is heavily environmentally-motivated. Additionally, further analysis of the sign type categories as created by @dutkiewicz2020, as it is possible, when it comes to the simpler signs, for identical creation to occur independently.

However, this preliminary research has been able to produce some conclusions to be reasonably confident in. Firstly, given that we know definitively that distribution and variation of sign types is not the result of geographic distance, we can conclude that the data is likely relevant to understanding patterns of cultural diversity. Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups. These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic Homo sapiens in Europe.

# Conclusion


