---
title: "Changes in Groupings of Geometric Signs on Paleolithic Mobile Objects as a Model of Social Dynamics Across Aurignacian Time Periods"
author: Sophie Cooper
format: html
echo: false
execute: 
  warning: false
bibliography: references.bib
zotero: "Signbase-study"
link-citations: true
---

# Introduction

The Upper Paleolithic period experienced a rise in more complex social organizations across Europe [@Boyd_Silk_2020] coupled with biological evolution. Another distinctive quality of the Upper Paleolithic is the large rise in artwork, both parietal and mobile. Archaeologists have frequently relied upon that artwork as a proxy for cultural difference in order to understand the shifting patterns of social organization and diversity of the Upper Paleolithic [@baker2024evidence, @Sauvet_2018, @Kuhn_Stiner_2007, @Vanhaeren_d]. Considered to be the starting point of the Upper Paleolithic, the Aurignacian technocomplex (ca. 43 - 30 k BP) is commonly linked to the dispersal of Homo sapiens across Western Eurasia, and is characterized by lithic and osseus markers [@Tejero_Grimaldi_2015; @Tartar_2012], some of which are decorated with geometric signs [@dutkiewicz2020]. In this paper, we investugate the social organization dynamics by analyzing the distribution and variation of geometric sign types found on mobile artefacts from the Upper Paleolithic in Europe across four phases of the Aurignacian period.

Social organization dynamics of the Upper Paleolithic have previously been explored using statistical analyses of artifacts to understand patterns of cultural groups. Multivariate analyses of beads from the Gravettian technocomplex suggests an east to west cline of nine cultural groups [@baker2024evidence]. Seriation analysis, PCoA, and network analysis were used to identify the groups from 134 discrete bead types recovered from both burial and occupation sites. To validate their claims of cultural groups, @baker2024evidence used a Mantel test to evaluate an isolation-by-distance hypothesis, which proposes that cultural difference can be primarily explained by geographic distance, finding that geographic distance alone did not solely account for the bead distribution. The nine geographically discrete groups encompass the regions of eastern, northwestern and central Europe, the northern and southern Iberian peninsula, southern and northern Italy, and the eastern and western Mediterranean regions.

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. They analyzed 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East. Geographically cohesive groups sharing similar ornament type associations were identified by seriation, correspondence, and GIS analyses. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. @Vanhaeren_d interpret the variation in personal ornament types as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.

Cultural difference has also been identified from geometric signs in Upper Paleolithic parietal art from the Cantabrian region of Spain [@Sauvet_2018]. Forty-two quadrilateral geometric signs from four cave sites, likely associated with the Magdalenian period, spanning a distance of 30 kilometers were recorded. Multivariate analysis of 45 morphological characteristics of the signs revealed two clusters. To interpret these clusters, @Sauvet_2018 draw on case studies of geometric signs as identity markers, including South African engraved ostrich shells from 60 k BP, maker's marks on Gallo-Roman pottery, and 19th century French craft guild members' personal marks. Using these case studies as analogies, @Sauvet_2018 interpret the Magdalenian geometric signs as identity markers, and the two clusters as representative of cultural groups.

In this study, we extend the use of geometric signs as identity markers to investigate cultural groups in the Aurignacian period. The Aurignacian period spans roughly 43 - 30 k BP, and is largely characterized by material culture indicators, such as carinated scrapers, small-flake tools and split-based bone and ivory tools, as well as an ongoing evolutionary shift from Neanderthals to early modern humans [@Chu_Richter_2020]. The Aurignacian can be divided into phases, again mainly distinguished by shifts in material culture; Proto, Early, Evolved, and Late Aurignacian [@Tartar_2015], as can be seen in @fig-aur-table. @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021 divided the Aurignacian into two main phases, Aur-P1 (Proto/Early), which corresponds with the time period of 43–37 k BP, and Aur-P2 (Evolved/Late), which corresponds with the time periods 37-32 k BP. @Banks_dErrico_Zilhao_2013 used Bayesian modeling methods to further divide Aur-P1 into the Proto and Early Aurignacian, placing the Proto-Aurignacian at 41.5 - 39.9 k BP, and the early Aurignacian at 39.8 - 37.8 k BP. They found that the Proto-Aurignacian occurs during the Greenland Interstadials (GI) 9 and 10, periods of climatic amelioration, while the Early Aurignacian began with the Henrich Stadial, which is characterized by dry and arid conditions, and ends with GI 8.

```{r}
#| label: fig-aur-table
#| fig-cap: "Table of data on Aurignacian periods [@Banks_dErrico_Zilhao_2013;  @Shao_Limberg_Klein_Wegener_Schmidt_Weniger_Hense_Rostami_2021; @Chu_Richter_2020; @Fletcher_Sánchez]"
library(knitr)
library(tidyverse)
library(readxl)

aur_table <- read_excel("data/aurignacian-periods.xlsx")

kable(aur_table)

```

We hypothesize that these periods, and the environmental changes across the Aurignacian, may correspond to changes in social dynamics of cultural groups, as represented by patterns in the distribution of geometric signs. We aim to address the following questions: How do grouping dynamics of geometric signs change across the four phases of the Aurignacian? How do the groupings vary in diversity, size, connectivity, and boundaries? Lastly, what do the patterns tell us about dynamics of social organization throughout the Aurignacian? We answer these questions by first identifying clusters within each time period using seriation and network analyses. We then test the strength of the clusters using modularity, perMANOVA, and PERMDISP2 tests. Lastly, we use Mantel tests to evaluate the null hypothesis of isolation-by-distance for each time period.

# Methods and Materials

```{r}
## clean the data
library(magrittr)
library(rcarbon)
signbase_full <- read_csv("data/signBase_Version1.0.csv")
nsites <- signbase_full %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts <- nrow(signbase_full)
ncountries <- signbase_full %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes <- ncol(signbase_full)

signbase_full_clean <- signbase_full %>%  # 511 rows
  filter(site_name != "Willendorf",
         site_name != "Riparo di Fontana Nuova",
         site_name != "Muralovka",
         site_name != "Shanidar Cave",
         site_name != "Hayonim Cave",
         site_name != "El Salitre",
         site_name != "Grotte De La Princesse Pauline",
         site_name != "Šandalja II") %>% 
  dplyr::select(-other, -rectangle) %>% 
  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>% 
  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))

signbase_years <- signbase_full_clean %>%
  drop_na(date_bp_max_min) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
  separate(date_bp_max_min,
           sep = " - ",
           c("date_bp_max", "date_bp_min"),
           remove = FALSE) %>% 
  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
                              str_extract(date_bp_max, ".*?(?=\\/)"),
                              date_bp_max)) %>% 
  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
                              str_extract(date_bp_min, ".*?(?=\\/)"),
                              date_bp_min)) %>% 
  separate(date_bp_max,
           sep = "±",
           c("date_bp_max_age",
             "date_bp_max_error")) %>% 
  separate(date_bp_min,
           sep = "±",
           c("date_bp_min_age",
             "date_bp_min_error")) %>% 
  drop_na(date_bp_max_age,
          date_bp_max_error) %>% 
  mutate(date_bp_max_age = parse_number(date_bp_max_age),
         date_bp_max_error = parse_number(date_bp_max_error))

signbase_years_cal <- 
  rcarbon::calibrate(signbase_years$date_bp_max_age,
                     signbase_years$date_bp_max_error,
                     verbose = FALSE) %>% 
  summary() %>% 
  data_frame()

signbase_years$MedianBP <- signbase_years_cal$MedianBP


signbase_full_clean <- signbase_full_clean %>% 
  inner_join(signbase_years %>% 
              dplyr::select(object_id, MedianBP))

signbase_full_clean <- signbase_full_clean %>%
  mutate(time_period = case_when((MedianBP > 41499) ~ "transitional",
                                 (MedianBP > 39799) ~ "proto_aurignacian",
                                 (MedianBP > 37799) ~ "early_aurignacian",
                                 (MedianBP > 31999) ~ "evolved_aurignacian"))

signbase_full_clean <- signbase_full_clean %>% 
  mutate(sign_total = rowSums((signbase_full_clean [,25:53]))) %>% 
  filter(sign_total > 0)


nsites_clean <- signbase_full_clean %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()
nartifacts_clean <- nrow(signbase_full_clean)
ncountries_clean <- signbase_full_clean %>% 
  pull(country) %>% 
  unique() %>% 
  length()
nsigntypes_clean <- nsigntypes - 1

pretty_print_dates <- 
  function(x) formatC(x, 
                      decimal.mark=".", 
                      big.mark=",", 
                      digits = 0, 
                      format = "f")
date_min <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  min()
date_max <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  max()
date_range <- date_max - date_min
date_min_pretty <- pretty_print_dates(date_min)
date_max_pretty <- pretty_print_dates(date_max)
date_range_pretty <- pretty_print_dates(date_range)
```

```{r}

lat_long_df <- signbase_full_clean %>% 
  dplyr::select(site_name, longitude, latitude) %>% 
  distinct(site_name, .keep_all = TRUE)

trans_unique_data <- signbase_full_clean %>% 
  filter(time_period == "transitional") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "transitional")

proto_unique_data <- signbase_full_clean %>% 
  filter(time_period == "proto_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "proto_aurignacian")

early_unique_data <- signbase_full_clean %>% 
  filter(time_period == "early_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "early_aurignacian")

evolved_unique_data <- signbase_full_clean %>% 
  filter(time_period == "evolved_aurignacian") %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(lat_long_df) %>% 
  mutate(time_period = "evolved_aurignacian")


trans_artifact_data <- trans_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)


proto_artifact_data <- proto_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)

early_artifact_data <- early_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)

evolved_artifact_data <- evolved_unique_data %>% 
  column_to_rownames("site_name") %>% 
  dplyr::select(line:star)


trans_unique_data <- trans_unique_data %>% 
  mutate(group = case_when((site_name == "Pod Hradem"|
                              site_name == "Fumane" |
                              site_name == "Labeko Koba") ~ "1",
                           (site_name == "El Castillo" | 
                              site_name == "Hohle Fels" | 
                              site_name == "Geissenklösterle") ~ "2"))

proto_unique_data <- proto_unique_data %>% 
  mutate(group = case_when((site_name == "Gatzarria" | 
                              site_name == "Hohlenstein-Stadel" | 
                              site_name == "Mladeč" | 
                              site_name == "Abri Pataud") ~ "1",
                           (site_name == "Geissenklösterle" |
                              site_name == "Grotte du Renne" |
                              site_name == "Vogelherd" |
                              site_name == "Spy" |
                              site_name == "La Ferrassie" |
                              site_name == "Les Cottés") ~ "2"))

early_unique_data <- early_unique_data %>% 
  mutate(group = case_when((site_name == "Grottes de Fonds-de-Forêt" |
                              site_name == "Riparo Bombrini"|
                              site_name == "Grotte de la Verpillière I"|
                              site_name == "Vindija Cave" |
                              site_name == "Trou al'Wesse") ~ "1",
                           (site_name == "Hohle Fels" | 
                              site_name == "Solutré" |
                              site_name == "Castanet"|
                              site_name == "Cellier" | 
                              site_name == "Blanchard") ~ "2"))


evolved_unique_data <- evolved_unique_data %>% 
  mutate(group = case_when((site_name == "Les Rois" |
                              site_name == "Gargas" |
                              site_name == "La Viña"|
                              site_name ==  "Sirgenstein Cave") ~ "1",
                           (site_name == "Bockstein-Törle"|
                              site_name == "Vogelherd" |
                              site_name == "Hohle Fels"|
                              site_name == "Trou Magrite"|
                              site_name == "Vindija Cave") ~ "2"))



time_unique_data_full <-rbind(trans_unique_data, proto_unique_data, 
                              early_unique_data, evolved_unique_data)

time_unique_data_full$time_period <- factor(time_unique_data_full$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("transitional", "proto-aurignacian",
                                                       "early aurignacian", "evolved aurignacian"))



```

```{r}
# mantel abundance
library(vegan)

set.seed(200)

artifact_data <- time_unique_data_full %>% 
  column_to_rownames("id") %>% 
  select(line:star)

euc_data <- artifact_data %>% 
  mutate(sign_count = rowSums(.)) %>% 
  dplyr::select(sign_count)

  
euc_dist <- dist(x = euc_data, method = "euclidean")
jac_full <- vegan::vegdist(artifact_data, method = "jaccard", binary = TRUE)

euc_mantel <- vegan::mantel(euc_dist, jac_full, permutations = 1000)
p_value_abundance_correlation <- round(euc_mantel$signif, 3)
r_value_abundance_correlation <- round(euc_mantel$statistic, 3)
```

## Data

```{r fig.height= 10, fig.width=10}
#| label: fig-site-map
#| fig-cap: "Graph of artifact count by time period and map of sites, divided by time period"

library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)
library(ggpubr)
library(cowplot)

signbase_full_clean$time_period <- factor(signbase_full_clean$time_period, 
                                            levels = c("transitional", "proto_aurignacian", 
                                                       "early_aurignacian", "evolved_aurignacian"), 
                                            labels = c("transitional", "proto-aurignacian",
                                                       "early aurignacian", "evolved aurignacian"))
  

date_plot <- ggplot(signbase_full_clean) +
  aes(x = MedianBP, fill = time_period) +
  geom_bar(width = 150) +
  scale_x_reverse() +
  labs(fill = "Time period") +
  theme(legend.key.size = unit(0.5, "cm"))
  

signbase_sf <- time_unique_data_full  %>% 
  st_as_sf(coords = c("longitude", "latitude"),
           remove = FALSE,
           crs = 4326)

world <- ne_countries(scale = "medium", returnclass = "sf")

Europe <- world[which(world$continent == "Europe"),]

site_map <- ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2) +
  theme_minimal() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank())
  
plot_grid(date_plot, site_map, nrow = 2, align = "hv", rel_heights = c(1, 2))
```

Our data for this paper comes from SignBase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, which was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation [@dutkiewicz2020]. The creators of the database identified distinct geometric sign types, and recorded their occurrences on mobile artefacts from the Paleolithic. Identification and categorization of the sign types had already been performed by the creators of the SignBase database, the methods for which are described in @dutkiewicz2020, and we adhered to those categories and placements.

The data set consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries, and records `r nsigntypes` geometric sign type found on these artifacts. We excluded signs if they fell under the "other" category, as the wide range of possibilities that could qualify as "other" means that it would be difficult to establish a definitive connection. We also excluded all data from Willendorf (only one sign present), and Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave (due to their extreme geographic distance from the other sites). Lastly, data from Grotte De La Princesse Pauline and Šandalja II was excluded, due to being dated much earlier than the rest of the artifacts.

These exclusions resulted in `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types. All our tests were run with presence/absence data, in order to keep the data from being affected by outliers like Vogelherd. We ran a Mantel test between an abundance distance matrix and and a Euclidean distance matrix in order to make sure that the exclusion of abundance would not significantly impact our results [@Lycett_2019], and found that it was not quite statistically significant (P = `r p_value_abundance_correlation`, Mantel R = `r r_value_abundance_correlation`).

Maximum and minimum BP dates were provided by Signbase, and calibration was done through the rcarbon package in RStudio [@Crema_Bevan_2021]. The calibrated radiocarbon dates of the artifacts range from approximately `r date_min_pretty` BP to `r date_max_pretty` BP, spanning an overall time range of `r date_range_pretty` years. We divided the data up into four periods in accordance with the date ranges provided in @fig-aur-table.

## Testing for Cultural Groupings

Seriation analysis is the arrangement of data into a linear order to reveal patterns, and is a common technique in analyzing sets of archaeological data [@Hahsler_Hornik_Buchta]. Seriation in this paper was performed using code designed to mimic the Browner-Kile seriation algorithm [@Brower_Kile_1988], in which the data is presented in a matrix diagram, and a unidimensional sequence is generated by reordering the rows and columns to group the presences along the diagonal. This is done so by calculating the mean row and column positions of presences, and arranging the rows and columns by respectively.

Network analysis represents the relationship between nodes [@Mills_2017]. In this network analysis, each node is a site, and each edge is a measure of sign similarity, with the direct value represented by edge weight. The results were plotted with the Fruchterman-Reingold layout [@Fruchterman_Reingold_1991], a force-directed graph layout for uniform edge lengths.

## Testing Strength of Cultural Groupings

The modularity test quantifies the randomness of the groupings by measuring the fraction of the edges in the network that fall within groups, minus the expected value if the edges were placed at random [@Newman_2006, @Newman_Girvan_2004]. The results are expressed in a score of 0 to 1, with 0 meaning no significant subgroups, and 1 meaning perfect groupings. The PerMANOVA test quantifies the level of variation between groups versus the level of variation within groups [@Anderson_2017], with the results expressed by an R-statistic, and a P-score.

PERMDISP is a multivariate extension of Levene's test [@Anderson_2006, @Anderson_Ellingsen_McArdle_2006], for the analysis of multivariate homogeneity of group dispersion. In this study, we perform the PERMDISP test upon the Jaccard matrix as a measure of compositional dissimilarity. We did so using the "betadisper" function from the RStudio package 'vegan', which mimics the PERMDISP2 procedure [@Vegan2001]. In this code, the non-euclidean distance matrix is reduced to principal coordinates to embed them within a Euclidean space. The results are used to produce a visual of the level of overlap between the groups in terms of sign content.

## Testing Isolation-by-Distance Hypothesis

The isolation-by-distance framework proposes that cultural difference can be primarily explained by geographic distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Paleolithic early modern humans that are dispersed according to geography. Mantel tests are used here for each time period to determine the correlation between sign distribution and geography. The Mantel test is run between a geographic distance matrix and a Jaccard dissimilarity matrix. The test seeks to determine the level of correspondence between the matrices through permuational evaluation of the null distribution [@Mantel_Smouse], producing a Mantel-R score and a p-score as a result.

# Results

## Testing for Cultural Groupings

```{r}
library(ggpubr)
library(phangorn)
library(ggraph)
library(statnet)
library(ecodist)
library(ape)
library(ggrepel)
library(cowplot)
library(igraph)

mplot <- function(x, ..., fill_colour = "black", title = NULL, row_labels = NULL, col_labels = NULL, country_colors = NULL) {
  d <- data.frame(
    x = rep(1:ncol(x), each = nrow(x)),
    y = rep(nrow(x):1, ncol(x)),
    z = as.integer(x)
  )
  
  if (!is.null(country_colors)) {
    d$country_color <- rep(country_colors, ncol(x))
  }
  
  if (is.null(col_labels)) col_labels <- colnames(x)
  if (is.null(row_labels)) row_labels <- rev(rownames(x))
  
  p <- ggplot(d, aes(x, y)) +
    geom_tile(col = "grey50", linewidth = 0.5, aes(fill = ifelse(z == 1, country_color, "white"))) +
    scale_fill_identity() +  
    coord_equal(expand = FALSE) +
    scale_x_continuous(
      breaks = 1:ncol(x),
      labels = col_labels,
      position = "top"
    ) +
    scale_y_continuous(
      breaks = 1:nrow(x),
      labels = row_labels
    ) +
    theme(
      axis.text = element_text(size = rel(1)),
      axis.ticks = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 1)
    ) +
    labs(x = "", y = "") +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)
  
  p
}

increase_focus <- function(x, country = NULL) {
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  row_order <- order(mcp, na.last = TRUE)
  x <- x[row_order, ]
  
  if (!is.null(country)) {
    country <- country[row_order]
  }
  
  mrp <- apply(
    x, 
    MARGIN = 2, 
    FUN = \(z) { mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  x <- x[, order(mrp, na.last = TRUE)]
  
  list(matrix = x, country = country)
}

concentrate <- function(x, country = NULL, max_iter = 100) {
  old <- x
  not_identical <- TRUE
  iter <- 0
  
  while (not_identical && iter < max_iter) {
    result <- increase_focus(old, country)
    new <- result$matrix
    country <- result$country
    not_identical <- !identical(old, new)
    old <- new
    iter <- iter + 1
  }
  
  if (iter == max_iter) {
    warning("Reached maximum iterations without convergence.")
  }
  
  list(matrix = new, country = country)
}

produce_clusters <- function(artifact_data, artifact_data_unique, method = "seriation"){
  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0)
  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
  dm <- as.matrix(jac)
  
  if(method == "seriation"){
  artifact_matrix <- as.matrix(artifact_data %>% 
                                 mutate(across(everything(), ~replace(., . > 1, 1))))
  rownames(artifact_matrix) <- rownames(artifact_data)
  
  x_df <- artifact_data_unique %>% 
    left_join(signbase_full_clean,
              by = "site_name")
  countries <- x_df$group
  unique_countries <- unique(countries)
  color_palette <-  RColorBrewer::brewer.pal(length(unique_countries), "Set3")
  row_colors <- setNames(color_palette, unique_countries)
  result <- concentrate(artifact_matrix, country = countries)
  concentrated_matrix <- result$matrix
  reordered_countries <- result$country
  countries_colors <- row_colors[countries]
  country_colors <- row_colors[reordered_countries]

  final_plot <- mplot(concentrated_matrix,
    country_colors = country_colors)}
  
  if(method == "network"){
  disim <- 1 - dm
  disim[disim<0.5] <- 0
  disim_net <- network(disim,
                       directed=F,
                       ignore.eval=F,
                       names.eval='weight')
  disim_net %v% 'vertex.names' <- row.names(artifact_data)
  set.seed(500)
  
  final_plot <- ggraph(disim_net, layout="fr") +
      geom_edge_link(aes(width = weight, 
                         alpha=weight), 
                     edge_colour = "black", 
                     show.legend= F,
                    ) +
      scale_edge_width(range=c(0.5,1.5)) +
      scale_edge_colour_gradient(low = "#CCCCCC",
                                 high = "#000000") +
      geom_node_point(alpha=1, size=5) +
      geom_node_text(aes(label=row.names(artifact_data)), 
                     size=3, 
                     family= "Arial", 
                     repel=T, max.overlaps = 100) +
      theme(text= element_text(size = 15), 
            panel.background = element_rect(fill = 'white')) +
    border(color = "black", size = 0.5)
    
  }
  return(final_plot)
}


strength_function <- function(artifact_data, group_data, type = "table"){
  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0)
  
  set.seed(500)
  jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
  dm <- as.matrix(jac)
  
  if(type == "table"){
  modul_matrix <-
    graph_from_adjacency_matrix(1 - dm, mode = "undirected")
  mem <- as_membership(parse_number(group_data$group))
  as_group <- modularity(modul_matrix, membership = mem)
  
  perman <- adonis2(jac~as.factor(group_data$group), 
                    method = "jaccard",
                    sqrt.dist = TRUE)
  perman <- perman %>% 
    data_frame() %>% 
    slice_head()
  perman_r <- perman$R2
  perman_p <- perman$`Pr(>F)`
  
  site_distances <- dist(cbind(group_data$longitude, group_data$latitude))
  mantel_distance <- vegan::mantel(site_distances, jac, permutations = 1000)
  p_value_distance_correlation <- round(mantel_distance$signif, 3)
  r_value_distance_correlation <- round(mantel_distance$statistic, 3)
  
  final_table <- data_frame("modularity" = as_group, 
                            "permanova r" = perman_r,
                            "permanova p" = perman_p,
                            "mantel r" = r_value_distance_correlation, 
                            "mantel p" = p_value_distance_correlation)

  return(final_table)}
  
  if(type == "dispersion"){
    set.seed(500)
    dispersion <- betadisper(jac, group = group_data$group)
    disp_plot<- plot(dispersion, hull = FALSE, ellipse=TRUE)
    return(disp_plot)}
}


trans_network <- function(artifact_data){
  artifact_data <- artifact_data %>% 
    select_if(~ !is.numeric(.) || sum(.) != 0)
    jac <- vegdist(artifact_data, "jaccard", binary = TRUE)
    dm <- as.matrix(jac)
    disim <- 1 - dm
    disim[disim<0.1] <- 0
    disim_net <- network(disim,
                         directed=F,
                         ignore.eval=F,
                         names.eval='weight')
    disim_net %v% 'vertex.names' <- row.names(artifact_data)
    set.seed(500)
    final_plot <- ggraph(disim_net, layout="fr") +
      geom_edge_link(aes(width = weight, 
                         alpha=weight), 
                       edge_colour = "black", 
                       show.legend=F) +
      scale_edge_width(range=c(0.5,1.5)) +
      scale_edge_colour_gradient(low = "#CCCCCC",
                                   high = "#000000") +
      geom_node_point(alpha=1, size=5) +
      geom_node_text(aes(label=row.names(artifact_data)), 
                     size=3, 
                     family= "Arial", 
                     repel=T, max.overlaps = 100) +
      theme(text= element_text(size = 15), 
            panel.background = element_rect(fill = 'white')) +
      border(color = "black", size = 0.5)
    return(final_plot)
    }

## mantel function tells us whether changes in sign makeup is correlated with chosen variable change. 
## Permanova tells us how similar in chosen variable members of group are

perm_function <- function(period = "none", variable = PCA, method = "permanova", signbase_sf = signbase_sf){
  set.seed(500)
  
  if(period == "none"){perm_df <- signbase_sf}
  
  else{perm_df <- signbase_sf %>% 
    filter(time_period == {{period}})}

  if(method == "permanova"){
  var_jac <- perm_df %>% 
      st_drop_geometry() %>% 
      column_to_rownames("id") %>% 
      dplyr::select({{variable}}) %>% 
      vegdist("jaccard")
    
  result <- adonis2(var_jac~as.factor(perm_df$group), 
                         method = "jaccard",
                         sqrt.dist = TRUE)}
  
  if(method == "mantel"){
    sign_jac <- perm_df %>% 
      st_drop_geometry() %>% 
      column_to_rownames("id") %>% 
      dplyr::select(line:star) %>% 
      vegdist("jaccard", binary = TRUE)
      
    var_sim <- perm_df %>% 
      st_drop_geometry() %>% 
      column_to_rownames("id") %>% 
      dplyr::select({{variable}}) %>% 
      dist()
    
    result <- vegan::mantel(sign_jac,  var_sim, permutations = 1000)
  }
  
  return(result)
}
```

### Seriation

```{r fig.height= 9, fig.width= 13}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"

seriation_trans <- produce_clusters(trans_artifact_data, trans_unique_data, method = "seriation")
seriation_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "seriation")
seriation_early <- produce_clusters(early_artifact_data, early_unique_data, method = "seriation")
seriation_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "seriation")

plot_grid(seriation_trans,seriation_proto, seriation_early , seriation_evolved, 
          ncol = 2, align = "hv", 
          labels = c("Transitional", "Proto", "Early", "Evolved"), label_size = 15, 
          scale = c(0.85, 1, 1, 1))


```

### Network Analysis

```{r fig.height= 5.5, fig.width= 8}
#| label: fig-neighbor-net
#| fig-cap: "Network analysis of the data"
network_trans <- trans_network(trans_artifact_data)
network_proto <- produce_clusters(proto_artifact_data, proto_unique_data, method = "network")
network_early <- produce_clusters(early_artifact_data, early_unique_data, method = "network")
network_evolved <- produce_clusters(evolved_artifact_data, evolved_unique_data, method = "network")

plot_grid(network_trans, network_proto, network_early, network_evolved, 
          ncol = 2, align = "hv", labels = c("Transitional", "Proto", "Early", "Evolved"), 
          label_size = 7, label_colour = "black", label_y = 1.02, label_x = -0.01)
```

Combining the results of the above tests allowed us to find two main groups of sites within each time period, as shown in @fig-group-table. The geographic distributions of the groups are shown in @fig-group-map, and the sign makeup of each group is shown in @fig-group-sign.

## Groups

```{r}
#| label: fig-group-table
#| fig-cap: "Table of the sites by group"
group_table <- read_excel("data/site-groups.xlsx")

kable(group_table)

```

```{r fig.height= 7, fig.width=10}
#| label: fig-group-map
#| fig-cap: "Map of sites faceted by time period and colored by group number"
  
ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf, aes(color = group)) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 30,
                  size = 2) +
  theme_minimal() +
  facet_wrap(~time_period) +
  theme(axis.text = element_blank(),
        axis.title = element_blank()) 

```

```{r fig.height= 10, fig.width=10}
#| label: fig-group-sign
#| fig-cap: "Bar plot of signs by time period, colored by group"

full_data_long <- signbase_sf %>% 
  pivot_longer(cols = line:star) %>% 
  filter(value != 0)

ggplot(full_data_long) +
  aes(x = name, fill = group) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) + 
  facet_wrap(~time_period, nrow = 4)

```

## Testing Strength of Cultural Groupings

```{r}
##Modularity test
trans_strength <- strength_function(trans_artifact_data, trans_unique_data, type = "table")
proto_strength <- strength_function(proto_artifact_data, proto_unique_data, type = "table")
early_strength <- strength_function(early_artifact_data, early_unique_data, type = "table")
evolved_strength <- strength_function(evolved_artifact_data, evolved_unique_data, type = "table")

full_table <- rbind(trans_strength, proto_strength, early_strength, evolved_strength) %>% 
  mutate(`time period` = c("transitional", "proto", "early", "evolved"), .before = 1)

modularity_table <- full_table %>% 
  dplyr::select(`time period`, modularity)

kable(modularity_table)
```

```{r}
##permanova test
permanova_table <- full_table %>% 
  dplyr::select(`time period`, `permanova r`, `permanova p`)

kable(permanova_table)
```

The R value from the perMANOVA tests tell us that for each of the time periods, around 20% of the variation in sign content can be explained by group membership, with the proto-aurignacian having the highest score of 27%, and the early Aurignacian having the lowest score of 21%. The perMANOVA p value quantifies the randomness of the groupings, with 0.05 as the threshold for definitive non-randomness. The only time period to not meet that threshold is the transitional period, with a p-score of 0.3.

```{r fig.height= 8, fig.width=5}
#| label: fig-perm-disp
#| fig-cap: "Multivariate homogeneity of group dispersions"

trans_strength_plot <- strength_function(trans_artifact_data, trans_unique_data, type = "dispersion")
proto_strength_plot <- strength_function(proto_artifact_data, proto_unique_data, type = "dispersion")
early_strength_plot <- strength_function(early_artifact_data, early_unique_data, type = "dispersion")
evolved_strength_plot <- strength_function(evolved_artifact_data, evolved_unique_data, type = "dispersion")



```

## Testing Isolation-by-Distance Hypothesis

```{r}
## Mantel test for geography vs signs
mantel_table <- full_table %>% 
  dplyr::select(`time period`, `mantel r`, `mantel p`)

kable(mantel_table)

```

The relevance of the isolation-by-distance hypothesis also varies from period to period.

# Discussion

In this paper, we set out to ask what differences in the patterns of groupings among geometric sign types found on mobile objects between phases of the Aurignacian could tell us about the social dynamics of that period.

One explanation for the sign type distribution could be that the groupings of the data represent cultural groups, as indicated by sign type patterns, and the geographic distribution between sites represent group mobility. @Whallon_2006 proposes a framework for understanding hunter-gatherer mobility, arguing that mobility occurs among hunter-gatherers for multiple reasons. The first two reasons, referred to by Whallon as logistical and residential mobility, occur when groups move in order to procure a new supply of resources once the more immediate ones have been depleted. Another type of mobility is network mobility, in which groups, or subsets of groups, move in order to make contact with other groups. This occurs in order to form a "safety net", in times of scarce resources, so that groups are assured of a positive reception if they are forced to move. Lastly, and similarly, is informational mobility, in which groups, or individuals, move with the primary goal of gathering information, typically about the environmental conditions and resources in other areas. Similarly, this operates as a "safety net" of sorts, in that these groups then posses knowledge on where to move in times of resource scarcity. One point of note is that, according to @Whallon_2006, decorative objects may be exchanged as a method of network formation, which may account for the presence of some signs at some sites.

Further research will need to be done on this topic in order to understand these patterns. In particular, the ecological and biogeographical conditions at the sites will need to be analyzed for potential connections, as mobility is heavily environmentally-motivated. Additionally, further analysis of the sign type categories as created by @dutkiewicz2020, as it is possible, when it comes to the simpler signs, for identical creation to occur independently.

However, this preliminary research has been able to produce some conclusions to be reasonably confident in. Firstly, given that we know definitively that distribution and variation of sign types is not the result of geographic distance, we can conclude that the data is likely relevant to understanding patterns of cultural diversity. Secondly, we have found clusters of sites that correspond in sign type, possibly indicating mobility of specific cultural groups. These conclusions contribute to, and broaden our current understanding of cultural and biological diversity among Upper Paleolithic Homo sapiens in Europe.

# Conclusion

# References
