---
title: "Preliminary Report"
format: html
execute: 
  warning: false
bibliography: references.bib
---

## Intro

Signbase, an open-access database collection of geometric signs on mobile objects from Paleolithic-era Europe, was launched in 2020, with the explicit goal being to enable quantitative analyses of abstract geographical representation (Dutkiewicz, et al., 2020). The creators of the database identified around 30 distinct geometric sign types, and recorded their occurances on mobile artefacts from the Paleolithic. In this paper, we will analyze the data from Signbase to see if we can find evidence of patterns indicating distinct cultural groups. [@dutkiewicz2020; @baker2024evidence]

## Background

Our understanding of the patterns of cultural and genetic diversity among the hominids of the Upper Paleolithic remains incomplete. The evidence-by-isolation framework proposes that similarity in cultural features will be primarily determined by geographical distance. Assuming that this framework is true, we would hypothetically see patterns of cultural features of Upper Palelolithic hominids that are dispersed according to geography. @petraglia1994

In this paper, we use geometric signs on mobile objects as an indicator of cultural variation. Baker, et al. write that "It is generally accepted that artefacts with exclusively symbolic functions, such as personal ornaments and mobiliary art, are more informative than functional objects when characterizing cultural and social systems as well as individuals’ role within society". \[summarise Baker et al, methods, materials, results\] Similarly, Vanhaeren, et al. argues for the advantages of studying personal ornamentation as an indicator for cultural groups. Some of the reasons that they point to are their exclusively symbolic function, and the well-documented history of societies using personal ornamentation as cultural indicators. \[summarise Vanhaeren et al, methods, materials, results\]

Our use of geometric signs, in particular, is also supported by Sauvet, et al., where geometric signs, and the variations between them, are used to identify cultural groups. They write, "It has been proposed at various times that these graphic expressions, generally called 'signs', constitute veritable 'identity markers' that can be perceived at multiple levels. Within a group, they serve to signal the membership of an individual to a social status or category. In intergroup relationships, they can have a more general role as markers of territory or ethnicity". \[summarise Sauvet et al, methods, materials, results\]

## Data

```{r}
##data
library(tidyverse)
library(magrittr)
library(rcarbon)
signbase_full <- read_csv("data/signBase_Version1.0.csv")

nsites <- signbase_full %>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()

nartifacts <- nrow(signbase_full)

ncountries <- signbase_full %>% 
  pull(country) %>% 
  unique() %>% 
  length()

nsigntypes <- ncol(signbase_full %>% 
                     select(line:other))

# BM: "full_clean" is a bit misleading here since we have dropped
# some sites and signs, you'll need to write about these steps in 
# your methods section 
signbase_full_clean <- signbase_full %>% 
  filter(site_name != "Willendorf",
         site_name != "Riparo di Fontana Nuova",
         site_name != "Muralovka",
         site_name != "Shanidar Cave",
         site_name != "Hayonim Cave",
         site_name != "El Salitre") %>% 
  select(-other, -rectangle)

nsites_clean <- signbase_full_clean%>% 
  pull(site_name ) %>% 
  unique() %>% 
  length()

nartifacts_clean <- nrow(signbase_full_clean)

ncountries_clean <- signbase_full_clean %>% 
  pull(country) %>% 
  unique() %>% 
  length()

nsigntypes_clean <- nsigntypes - 1

signbase_years <- signbase_full_clean %>% 
  drop_na(date_bp_max_min) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>% 
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>% 
  separate(date_bp_max_min,
           sep = " - ",
           c("date_bp_max", "date_bp_min"),
           remove = FALSE) %>% 
  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
                              str_extract(date_bp_max, ".*?(?=\\/)"),
                              date_bp_max)) %>% 
  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
                              str_extract(date_bp_min, ".*?(?=\\/)"),
                              date_bp_min)) %>% 
  separate(date_bp_max,
           sep = "±",
           c("date_bp_max_age",
             "date_bp_max_error")) %>% 
  separate(date_bp_min,
           sep = "±",
           c("date_bp_min_age",
             "date_bp_min_error")) %>% 
  drop_na(date_bp_max_age,
          date_bp_max_error) %>% 
  mutate(date_bp_max_age = parse_number(date_bp_max_age),
         date_bp_max_error = parse_number(date_bp_max_error))

signbase_years_cal <- 
  rcarbon::calibrate(signbase_years$date_bp_max_age,
            signbase_years$date_bp_max_error,
            verbose = FALSE) %>% 
  summary() %>% 
  data_frame()

date_min <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  min()

date_max <- signbase_years_cal %>% 
  pull(MedianBP) %>% 
  max()

date_range <- date_max - date_min
```

The data used was published in the Signbase database (https://www.signbase.org). It consists of `r nartifacts` artifacts found at `r nsites` sites in `r ncountries` countries. The database records `r nsigntypes` geometric sign type found on these artifacts. We decided to exclude any sign counts if they fell under the "other" category, as the wide range of possibilities that could qualify as "other" means that it would be difficult to establish a definitive connection. We also decided to exclude the data from five sites: Willendorf, Riparo di Fontana Nuova, Muralovka, Shanidar Cave, and Hayonim Cave. Willendorf was excluded because its only sign count fell under the "other" category, Otaslavice was excluded due to a lack of similarity to the other sites, and the rest were excluded due to geographic distance from the other sites. These decisions brought our data down to `r nartifacts_clean` artifacts found at `r nsites_clean` sites in `r ncountries_clean` countries, encapsulating `r nsigntypes_clean` sign types. The artifacts are determined to have been made in the Paleolithic era, specifically in the Aurignacien technocomplex. The calibrated radiocarbon dates of the artifacts range from approximately `r date_min` BP to `r date_max` BP, spanning an overall time range of `r date_range` years.

## Methods

We performed a seriation analysis on the data to identify groups. We also performed a PCoA, a Neighbour-Network analysis, and a PERMDISP2 on the data to visualize the grouping. The dissimilarity of the signbase data was represented using a Jaccard matrix in all of these methods. We performed permutation and modularity tests on the identified groups to determine how strong the grouping was. We also ran a mantel test using the Jaccard similarity matrix and a geographic distance matrix, to determine the influence of geography on the distribution of the signs. \[expand on the details for each method, and give a little justification, why did we use that method, and add citations\]

## Results

```{r}
##Add groups

groups <- list("1" = c("Abri Pataud", "El Rascaño", 
                     "Fumane", "La Viña",
                     "Lhotka", "Pod Hradem",
                     "Riparo Bombrini", "Sirgenstein Cave",
                     "Šandalja II"),
               "2" = c("Kvasice", "Labeko Koba", 
                        "Les Rois", "Slatinice",
                        "Göpfelsteinhöhle"),
               "3" = c("Aurignac", "d'Engihoul",
                       "Istállóskő Cave", "Hohlenstein-Stadel",
                       "Gargas", "Grotte de la Verpillière I", 
                       "Mladeč", "Brassempouy",
                       "Gatzarria"),
               "4" = c("Wildscheuer", "Breitenbach",
                        "Peskő Cave", "Vindija Cave",
                        "El Castillo"),
               "5" = c("Trou al'Wesse", "Maisières-Canal",
                        "Les Cottés", "Le Terme Pialat"),
               "6" = c("Tuto de Camalhot","Abri Lartet/Gorge d'Enfer",
                        "Castanet", "Le Trou Du Renard", 
                        "Grottes de Fonds-de-Forêt"),
               "7" = c("Nová Dědina", "Menton/Grottes du Grimaldi",
                        "Grotte du Renne", "Solutré", 
                        "Žlutava"),
               "8" = c("Trou Magrite", "Grotte De La Princesse Pauline",
                        "Grotte de Goyet", "Bockstein-Törle"),
               "9" = c("Vogelherd","Hohle Fels",
                        "Geissenklösterle", "Spy"),
               "10" = c("Cellier", "Blanchard", 
                         "La Ferrassie", "Hornos de la Pena",
                         "Otaslavice"),
               "11" = c("La Souquette", "Shelter Birów IV",
                         "Abri de Laussel", "Abri du Poisson"))
  
group_df <- 
  enframe(groups, 
          name = "group", 
          value = "site_name") %>%
  unnest(cols = site_name)

lat_long_df <- signbase_full_clean %>% 
  select(site_name, longitude, latitude) %>% 
  distinct(site_name, .keep_all = TRUE)

signbase_unique_groups <- signbase_full_clean %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum)) %>% 
  left_join(group_df) %>% 
  mutate(group = parse_number(group))

signbase_unique_groups <- signbase_unique_groups %>% 
  left_join(lat_long_df)

```

```{r}
##clean the data 
artifact_data <- signbase_unique_groups %>% 
  column_to_rownames(var = "site_name") %>% 
  select(line:star)

```


```{r}
#| label: fig-site-map
#| fig-cap: "Map of sites divided by group, where sign abundace is represented by point size"
##map
library(rnaturalearth)
library(rnaturalearthdata)
library(ggrepel)
library(sf)
abundance_data <- signbase_full_clean %>% 
  pivot_longer(cols= line:star,
               values_to = "sign_total") %>% 
  filter(sign_total != 0) %>% 
  mutate(longitude = as.character(longitude),
         latitude = as.character(latitude)) %>% 
  group_by(site_name) %>%
  summarize(across(where(is.numeric), sum))


abundance_data <- abundance_data %>% 
  left_join(lat_long_df) %>% 
  left_join(group_df) %>% 
  select(site_name, sign_total, longitude, latitude, group) %>% 
  mutate(group = parse_number(group))
  
signbase_sf <- 
  st_as_sf(abundance_data, 
           coords = c("longitude", "latitude"),
           remove = FALSE,
           crs = 4326)


world <- ne_countries(scale = "medium", returnclass = "sf")
Europe <- world[which(world$continent == "Europe"),]

ggplot(Europe) +
  geom_sf() +
  geom_sf(data = signbase_sf,
          aes(colour = as.factor(group),
              size = sign_total)) +
  coord_sf(xlim = c(-10,30), 
           ylim = c(35,53), 
           expand = FALSE) +
  geom_text_repel(data = signbase_sf,
                  aes(x = longitude ,
                      y = latitude,
                      label = site_name),
                  max.overlaps = 100,
                  size = 2) +
  theme_minimal() +
  facet_wrap(~group) +
  labs(color = "Group Number",
       size = "Sign Count")
```

```{r}
#| label: fig-seriation
#| fig-cap: "Seriation analysis of the data"

# BM: this seriation needs to come before the grouping code above, since the groupd were read of the seriation plot

## seriation
library(ggpubr)
artifact_matrix <- as.matrix(artifact_data %>% 
 mutate(across(everything(), ~replace(., . > 1, 1))))
rownames(artifact_matrix) <- rownames(artifact_data)


mplot <- function(x,...){
  d <- data.frame(x = rep(1:ncol(x), each = nrow(x)),
                  y = rep(nrow(x):1, ncol(x)),
                  z = as.integer(x))
  ggplot(d, aes(x, y, fill = factor(z))) +
    geom_tile(col = "grey50", linewidth = 0.5) +
    scale_fill_manual(values = c("0" = "white", "1" = "black")) +
    coord_equal(expand = FALSE) +
    scale_x_continuous(breaks = 1:ncol(x), 
                       labels = colnames(x),
                       position = "top") +
    scale_y_continuous(breaks = 1:nrow(x), 
                       labels = rev(rownames(x))) +
    theme(axis.text = element_text(size = rel(0.5)),
          axis.ticks = element_blank(),
          legend.position = "none",
          plot.title = element_text(hjust = 0.25)) +
    labs(x = "Sign Type",
         y = "Site Name",
         fill = "Count") +
    rotate_x_text(angle = 90, hjust = NULL, vjust = NULL)}


increase_focus <- function(x){
  mcp <- apply(
    x, 
    MARGIN = 1, 
    FUN = \(z){ mean(which(z == 1)) }, 
    simplify = FALSE
  ) |> unlist()
  x <- x[order(mcp), ]
  mrp <- apply(
    x, 
    MARGIN = 2, 
    FUN = \(z){ mean(which(z == 1)) }, 
    simplify = FALSE) |> unlist()
  x[, order(mrp)]}

concentrate <- function(x){
  old <- x
  not_identical <- TRUE
  while(not_identical){
    new <- increase_focus(old)
    not_identical <- !identical(old, new)
    old <- new}
  new}

artifact_matrix |> concentrate() |> mplot()

```

```{r}
##graphs showing sign makeup of each group

graph_data <- artifact_data %>% 
  concentrate() %>% 
  rownames_to_column("site_name") %>% 
  left_join(group_df)

sign_type_data <- graph_data %>% 
  pivot_longer(cols= notch:hashtag,
               names_to = "sign_type",
               values_to = "count") %>% 
  dplyr::filter(count != 0) %>% 
  mutate(group = parse_number(group))

x_order <- graph_data %>% 
  select(-site_name,
         -group) %>% 
  colnames() 

ggplot(sign_type_data) +
  aes(x = sign_type) +
  geom_bar() +
  rotate_x_text(angle = 90, hjust = NULL, vjust = NULL) +
  labs(x = x_order) +
  scale_x_discrete(limits = c(x_order)) +
  facet_wrap(~group, nrow = 11)
```

```{r}
#| label: fig-neighbor-net
#| fig-cap: "Neighbor-Net analysis of the data"
##network analysis
library(vegan)
library(phangorn)

## neighborNet
jac <- vegdist(artifact_data, "jaccard", binary = TRUE)

dm <- as.matrix(jac)
nnet <- neighborNet(dm)

par("mar" = rep(1, 4))

# Reorder the tips according to the seriation grouping
nnet$tip.label <- nnet$tip.label[match(group_df$site_name, nnet$tip.label)]  

# Plot the network without tip labels initially
plot(nnet, 
     show.tip.label = TRUE,
      cex = 0.5)


## try new neighbor net
library(ggraph)
library(tidygraph)
library(igraph)
library(statnet)
library(intergraph)
library(FastKNN)
library(ggpubr)
library(reshape2)
library(patchwork)
library(ggmap)
library(seriation)
library(purrr)
library(tnet)
library(argosfilter)

disim <- 1 - dm
disim[disim<0.5] <- 0
disim_net <- network(disim,
                    directed=F,
                    ignore.eval=F,
                    names.eval='weight')

disim_net %v% 'vertex.names' <- 
  row.names(artifact_data)

set.seed(500)

ggraph(disim_net, layout="fr") +
  geom_edge_link(aes(width = weight, 
                     alpha=weight), 
                 edge_colour = "black", 
                 show.legend=T) +
  scale_edge_width(range=c(0.5,1.5)) +
  scale_edge_colour_gradient(low = "#CCCCCC",
                             high = "#000000") +
  geom_node_point(alpha=1, size=5) +
  geom_node_text(aes(label=row.names(artifact_data)), 
                 size=6, 
                 family= "Arial", 
                 repel=T) +
  theme(text= element_text(size = 20), 
        panel.background = element_rect(fill = 'white'))

# BM try yet another

# Ensure all edge lengths are positive
nnet$edge.length <- pmax(nnet$edge.length, 1e-6)

# Create the graph object
g <- graph_from_edgelist(nnet$edge, directed = FALSE)
E(g)$weight <- nnet$edge.length

# Create a layout for the nodes
layout <- layout_with_fr(g)  # Fruchterman-Reingold layout

# Create a data frame for nodes (coordinates and labels)
nodes <- data.frame(
  x = layout[, 1],  # x-coordinates
  y = layout[, 2],  # y-coordinates
  label = c(nnet$tip.label, rep("internal", nnet$Nnode))
)

# Create a data frame for edges (start and end coordinates)
edges_df <- data.frame(
  from = nnet$edge[, 1],  # Start node of each edge
  to = nnet$edge[, 2],    # End node of each edge
  weight = nnet$edge.length  # Corresponding edge weights
)

# Create a data frame for the edge coordinates
edges_df$x_start <- nodes$x[edges_df$from]  # x-coordinate of start node
edges_df$y_start <- nodes$y[edges_df$from]  # y-coordinate of start node
edges_df$x_end <- nodes$x[edges_df$to]     # x-coordinate of end node
edges_df$y_end <- nodes$y[edges_df$to]     # y-coordinate of end node

# Plot the network
library(ggrepel)
ggplot() +
  geom_segment(data = edges_df, 
               aes(x = x_start, y = y_start, xend = x_end, yend = y_end),
               color = "grey", size = 1) +
  geom_point(data = nodes %>% 
               filter(label != "internal"),
             aes(x = x, y = y), color = "black", size = 3) +
  geom_text_repel(data = nodes %>% 
              filter(label != "internal"), 
            aes(x = x, y = y, label = label), 
            size = 3, vjust = -1) +
  theme_void() +
  theme(legend.position = "none")


```

```{r}
#| label: fig-pcoa
#| fig-cap: "PCoA analysis of the data"

##PCoA
library(ecodist)
library(ape)
pcoa <- pcoa(jac)

pcoa_df <- data.frame(pcoa1 = pcoa$vectors[,1], 
                      pcoa2 = pcoa$vectors[,2])

pcoa_df <- pcoa_df %>% 
  mutate(site_name = rownames(artifact_data))

library(ggrepel)
ggplot(pcoa_df) +
  aes(x=pcoa1, 
      y=pcoa2) +
  geom_point() +
  labs(x = "PC1",
       y = "PC2",) +
  theme(title = element_text(size = 10)) +
  geom_text_repel(aes(label = site_name), 
                  size = 2,
                  max.overlaps = 20)




```

```{r}
##PCoA Hirearchical clustering
library(factoextra)
library(FactoMineR)
hire_data <- artifact_data %>% 
   mutate(across(everything(), ~replace(., . > 1, 1)))
  
signbase_pca <- PCA(hire_data)
signbase_hcpc <- HCPC(signbase_pca, graph = FALSE)

##phylogenetic tree
fviz_dend(signbase_hcpc,
          cex = 0.7,
          color_labels_by_k = FALSE,
          type = "phylogenic")

##dendrogram
fviz_dend(signbase_hcpc,
          color_labels_by_k = FALSE,
          cex = 0.7)

## Cluster graph
fviz_cluster(signbase_hcpc,
             repel = TRUE,
             show.clust.cent = FALSE, 
             color_labels_by_k = FALSE,
             ellipse.alpha = 0,
             ggtheme = theme_bw(),
             main = "Factor map")

##to see clusters in a df
hcpc_dataframe <- signbase_hcpc$data.clust

## to see quantitative variables for each cluster
signbase_hcpc$desc.var$quanti

## to see principal dimensions associated with clusters
signbase_hcpc$desc.axes$quanti


```

```{r}
#| label: fig-perm-disp
#| fig-cap: "Multivariate homogeneity of group dispersions"
## permanova test
perman <- adonis2(jac~as.factor(signbase_unique_groups$group), 
                  method = "jaccard",
                  sqrt.dist = TRUE)

dispersion <- betadisper(jac, group = signbase_unique_groups$group)

disp <- (permutest(dispersion))

perman <- data.frame(perman) %>% 
  mutate(type = c("Model", "Residual", "Total")) %>% 
  filter(type == "Model")

disp <- disp$tab %>% 
  na.omit()

permu_table <- data.frame("Permutation_test_adonis" = c(perman$R2, perman$F, perman$Pr..F.),
           "Permutation_test_homogeneity" = c("NA", disp$F, disp$`Pr(>F)`))
rownames(permu_table) <- c("R-squared", "F", "P-score")

permu_table
plot(dispersion, hull = FALSE, ellipse=TRUE)

```

```{r}
##modularity test
library(igraph)
set.seed(500)

modul_matrix <-
  graph_from_adjacency_matrix(1 - dm, mode = "undirected")

gn <- cluster_edge_betweenness(modul_matrix)
wt <- cluster_walktrap(modul_matrix, steps = 4)
lv <- cluster_louvain(modul_matrix)


mem <- as_membership(signbase_unique_groups$group)

as_group <- modularity(modul_matrix, membership = mem)

create_tibble <- data.frame("Seriation groups" = c(as_group),
                            "Girvan-Newman" = c(max(gn$modularity)), 
                            "Walktrap" = c(max(wt$modularity)),
                            "Louvain" = c(lv$modularity))
print(create_tibble)


#experimenting with increasing modularity
core <- coreness(modul_matrix)
col_set <- heat.colors(max(core), rev = TRUE)
set.seed(2509)
plot(modul_matrix, vertex.color = col_set[core])

core_df <- core %>% 
  data_frame()
compo <- igraph::components(modul_matrix)
core_df$components <- compo$membership
colnames(core_df) <- c("coreness", "components")
rownames(core_df) <- signbase_unique_groups$site_name
core_df <- core_df %>% 
  rownames_to_column("site_name")

```

```{r}
# mantel tests
library(ape)
library(ade4)
library(Lock5Data)

set.seed(36)

site_distances <- dist(cbind(signbase_unique_groups$longitude, signbase_unique_groups$latitude))

rtest <- mantel.rtest(site_distances, jac, nrepet = 1000)

mantel <- vegan::mantel(site_distances, jac, permutations = 1000)

site_distances_matrix <- as.matrix(site_distances)

mantel_test <- mantel.test(site_distances_matrix, dm, nperm = 1000)


mantel_dataframe <- data_frame("p score" = c(rtest$pvalue, mantel$signif, mantel_test$p),
                               "Mantel Statistic R" = c(rtest$obs, mantel$statistic, NA))

rownames(mantel_dataframe) <- c("Mantel.rtest()", "Mantel()", "Mantel.test()")

mantel_dataframe

bur_man_cor <- mantel.correlog(jac, site_distances, n.class=50, break.pts=NULL, cutoff=FALSE, r.type="pearson", nperm=999, mult="holm", progressive=TRUE)

plot(bur_man_cor)

```
