---
title: "Supplementary Material: Site Phase Assignments"
subtitle: "Geometric signs as proxies for cultural connectivity during the Western Eurasian Aurignacian"
author:
  - Sophie Cooper:
      institute: uw
      orcid: xxx
  - Aleah Rosner:
      institute: uw
      orcid: xxx
  - Eva Oryn:
      institute: uw
      orcid: xxx
  - Kenneth B. Vernon:
      orcid: 0000-0003-0098-5092
      institute:
        - uu
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - uw: Department of Anthropology, University of Washington, Seattle, USA
  - uu: Scientific Computing and Computing Institute, University of Utah, Salt Lake City, USA
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 1000
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/elsevier-vancouver.csl" # Insert path for the bib-style
---

## Overview

This supplement documents the rationale and procedure for assigning archaeological sites to Aurignacian phases. Phase assignments in the main analysis are based on specialist technological attributions from the published literature, not solely on radiocarbon dates. This supplement summarises the evidence for each reclassification and provides the complete site-level status table.

Our phase assignments follow specialist technological attributions from the published literature, not radiocarbon dates alone, which can conflict with the archaeological content of a layer. Recent reassessments are therefore incorporated directly into our main analysis: Abri Pataud is placed in the Early Aurignacian (Aur-P1) on the basis of its lithic technology [@Chiotti_2005; @Higham_Abri_Pataud_2011], La Ferrassie in the Early Aurignacian (Aur-P1) on the basis of Layer 7 directly above the Châtelperronian [@Talamo_La_Ferrassie_2020], Mladeč in the Evolved Aurignacian (Aur-P2) on the basis of its massive-base bone points and ~31 ka ¹⁴C BP (~35–36 ka cal BP) dates [@Wild_et_al_2005], and the Swabian Jura site of Geissenklösterle in the Early Aurignacian [Aur-P1, @Bataille_Conard_2018; @Bataille_Conard_2019]. Hohle Fels is assigned at the layer level rather than as a single site: its lower Aurignacian horizons (AH V) are placed in the Early Aurignacian (Aur-P1), whereas its upper horizons (AH IV, III and the II-series) are placed in the Evolved/Late Aurignacian (Aur-P2), following the demonstration that the assemblages of AH IV (and the overlying IIIa and IIIb) are closely comparable to the Western European Late Aurignacian [@Dinnis_et_al_2019; @Bataille_Conard_2018; @Bataille_Conard_2019]. Two sites are excluded from the analysis because their assemblages are too mixed to assign to a single phase reliably: Spy [@Flas_et_al_2013; @Flas_2015] and El Castillo. The robustness of these assignments to reverting to purely date-based criteria is tested in S2 "Site Phase Assignments", where stratigraphic mixing concerns at Vogelherd are also addressed [@Schurch_Conard_2026_reassess]. We detail the rationale and procedure for assigning archaeological sites to Aurignacian phases further below.

## Data pipeline

The following code loads the raw SignBase dataset, parses radiocarbon dates, calibrates them, and assigns objects to phases. Sites excluded from the main analysis are filtered out, and sites reassigned based on technological attribution are overrides in the `case_when()`. Date parsing now handles both ends of date ranges, routes OSL ages around radiocarbon calibration, and uses IntCal20 explicitly. The default `date_variant = "midpoint"` is used here; sensitivity to variant choice is explored in Supplement S1.

```{r}
#| label: load-data
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(here)

signbase_full <- read.csv(here("data/signBase_Version1.0.csv"), header = TRUE)

# Source shared functions (includes clean_signbase with corrected date parsing)
source(here("paper/functions.R"))

signbase_full_clean <- clean_signbase(signbase_full, date_variant = "older")
```

```{r}
#| label: assign-phases
# NOTE: Phase assignment mirrors paper/functions.R clean_signbase case_when
# (single source of truth). Duplicated here only to document the S2 table
# derivation step-by-step; any change must be mirrored in functions.R.
# Main baseline uses date_variant="older" so the report-based Aur-P1 20 /
# Aur-P2 9 partition matches frozen manual_groups. Midpoint/younger are
# explored as sensitivity in S1.

# Re-apply the same technology overrides + date thresholds as in functions.R
# (S2 re-derives for documentation, but the object table already carries
# phase2 from clean_signbase).
signbase_full_clean <- signbase_full_clean %>%
  mutate(time_period = case_when(
    # --- Exclusions (mixed assemblages) ---
    site_name == "Spy"         ~ NA_character_,
    site_name == "El Castillo" ~ NA_character_,
    # --- Reclassifications based on technological attribution ---
    # Hohle Fels is assigned at the layer level: the lower Aurignacian horizons
    # (AH Vb, Vab, Vaa, Va) are Early Aurignacian (Aur-P1), while the upper
    # horizons (AH IV, IIIb, IIIa and the II-series) are Evolved/Late
    # Aurignacian (Aur-P2), following Dinnis et al. (2019) and Bataille &
    # Conard (2018). The single object from the mixed layer IIIa-V spans the
    # whole sequence and is excluded.
    site_name == "Hohle Fels" & layer %in% c("Va", "Vaa", "Vab", "Vb") ~ "early_aurignacian",
    site_name == "Hohle Fels" & layer %in% c("IV", "IIIa", "IIIb", "Iid", "Iida", "Iie") ~ "evolved_aurignacian",
    site_name == "Hohle Fels" & layer == "IIIa-V" ~ NA_character_,
    site_name == "Geissenklösterle"  ~ "early_aurignacian",
    site_name == "Vogelherd"         ~ "early_aurignacian",
    site_name == "La Ferrassie"      ~ "early_aurignacian",
    site_name == "Abri Pataud"       ~ "early_aurignacian",
    site_name == "Mladeč"            ~ "evolved_aurignacian",
    # --- Override: Gatzarria Cjn2 is Proto-Aurignacian ---
    site_name == "Gatzarria"         ~ "proto_aurignacian",
    # --- Default: date-based assignment (older variant for main) ---
    MedianBP > 39799 ~ "proto_aurignacian",
    MedianBP > 37799 ~ "early_aurignacian",
    MedianBP > 31999 ~ "evolved_aurignacian"
  )) %>%
  filter(!is.na(time_period)) %>%
  mutate(phase2 = ifelse(time_period %in% c("proto_aurignacian", "early_aurignacian"),
                         "Aur-P1", "Aur-P2")) %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  filter(sign_total > 0)
```

```{r}
#| label: summary-counts
site_summary <- signbase_full_clean %>%
  group_by(phase2) %>%
  summarise(
    sites = n_distinct(site_name),
    objects = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(objects))

# site_summary
```

```{r}
#| label: site-basis
# Literature-rationale text ONLY: the site names and object counts come from
# the pipeline above; only the "Basis" citation strings are recorded by hand.
# Any site not listed here is date-assigned (the default case_when branch).
site_basis <- data.frame(
  site_name = c("Hohle Fels", "Geissenklösterle", "Vogelherd", "La Ferrassie",
                "Abri Pataud", "Gatzarria", "Mladeč"),
  assignment = "technology",
  basis = c(
    "Layer-level split: AH V Early Aurignacian (Aur-P1); AH IV, III and II-series Evolved/Late Aurignacian (Aur-P2) [@Bataille_Conard_2018; @Bataille_Conard_2019; @Dinnis_et_al_2019]",
    "Early Aurignacian [@Bataille_Conard_2018]",
    "Early Aurignacian [@Schurch_Conard_2026_reassess; @Bataille_Conard_2018]",
    "Early Aurignacian [@Talamo_La_Ferrassie_2020]",
    "Early Aurignacian [@Chiotti_2005; @Higham_Abri_Pataud_2011]",
    "Proto-Aurignacian, layer Cjn2 [@Barshay_Szmidt_2012]",
    "Evolved Aurignacian [@Oliva_2006]"
  )
)

excluded_basis <- data.frame(
  site_name = c("Spy", "El Castillo"),
  assignment = "excluded",
  basis = c(
    "Mixed assemblage, no single Proto-Aurignacian attribution [@Flas_et_al_2013; @Flas_2015; @Semal_2009]",
    "Mixed/contaminated layers preclude reliable assignment"
  )
)

# Object counts per retained site, tagged by assignment basis
site_objects <- signbase_full_clean %>%
  group_by(site_name, phase2) %>%
  summarise(objects = n(), median_bp = median(MedianBP), .groups = "drop") %>%
  left_join(site_basis, by = "site_name") %>%
  mutate(assignment = ifelse(is.na(assignment), "date", assignment))

# Date-assigned sites get the threshold string generated from their own MedianBP
site_objects <- site_objects %>%
  mutate(basis = ifelse(
    assignment == "date",
    case_when(
      median_bp > 39799 ~ "MedianBP > 39,799",
      median_bp > 37799 ~ "MedianBP 37,799\u201339,799",
      TRUE               ~ "MedianBP 31,999\u201337,799"
    ),
    basis
  )) %>%
  mutate(basis = ifelse(
    site_name == "Fumane",
    "MedianBP > 39,799 (Protoaurignacian throughout sequence [@Falcucci_Peresani_2017; @falcucci2020breaking])",
    basis
  ))

knitr::kable(site_objects, caption = "Retained sites with object counts and assignment basis.")
```

```{r}
#| label: numeric-summary
# Inline-prose numbers computed from the pipeline (never hard-coded).
zero_sign_objects <- signbase_full %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  filter(sign_total == 0)

n_zero_objects <- nrow(zero_sign_objects)
n_zero_sites  <- n_distinct(zero_sign_objects$site_name)

n_retained_sites    <- n_distinct(signbase_full_clean$site_name)
n_retained_objects  <- nrow(signbase_full_clean)
n_retained_countries <- n_distinct(signbase_full_clean$country)

n_p1_sites   <- site_summary[["sites"]][site_summary$phase2 == "Aur-P1"]
n_p1_objects <- site_summary[["objects"]][site_summary$phase2 == "Aur-P1"]
n_p2_sites   <- site_summary[["sites"]][site_summary$phase2 == "Aur-P2"]
n_p2_objects <- site_summary[["objects"]][site_summary$phase2 == "Aur-P2"]
```

## Key literature findings

### Hohle Fels → layer-level assignment (AH V Early; AH IV, III, II-series Evolved)

Hohle Fels is the only site in the dataset whose internal sequence warrants splitting the phase assignment at the layer level rather than treating the whole site as a single phase.

- @Bataille_Conard_2018 : AH IV is a "techno-functional facies of the Swabian Aurignacian" with burin-core bladelet production; it shares features with the Early Aurignacian but is compositionally distinct from the underlying AH V assemblage, challenging the straightforward application of the Western European typo-chronological system to Central Europe.
- @Bataille_Conard_2019 : AH IIIa and IV share similar tool/core assemblages typical of the Swabian Aurignacian, forming a coherent upper unit that contrasts with AH V.
- @Dinnis_et_al_2019 : Hohle Fels IV shows "important similarities with the Western European Late Aurignacian"; its radiocarbon dates of 33–30 ka uncal BP are not older than those of the Western European Late Aurignacian. This is the direct basis for placing AH IV and the overlying III and II-series layers in Aur-P2 (Evolved/Late).
- AH V (Va, Vaa, Vab, Vb) is the earlier Aurignacian unit at Hohle Fels and is retained in Aur-P1 (Early Aurignacian); its calibrated median dates (~39.7–43.3 ka BP) are consistent with that placement.
- The single object recorded from the mixed layer "IIIa-V" spans the whole Aurignacian sequence and is therefore excluded from the analysis, in line with the treatment of the mixed-assemblage sites Spy and El Castillo.

### Geissenklösterle → Early Aurignacian (Aur-P1)

- @Bataille_Conard_2018 : The Geissenklösterle Aurignacian (AH III and II) is assigned to the Early Aurignacian on technological grounds.
- @Higham_2012 : Radiocarbon chronology of Geißenklösterle provides the direct dating basis for this assignment.
- Both Geissenklösterle layers remain in Aur-P1: the calibrated median dates of the SignBase radiocarbon determinations for AH II (~40.7 ka BP) and AH III (~42.7 ka BP) fall within the Aur-P1 range, so the technology- and date-based assignments agree for this site. A reassessment [places AH III later (~33–35 ka BP) [@zilhao2003aurignacian], but this conflicts with the site's own radiocarbon chronology and is not adopted here; the weight of the technological and stratigraphic evidence keeps both layers in Aur-P1.

### Vogelherd → Early Aurignacian (Aur-P1, site-level)

Vogelherd is kept as a single site-level Aur-P1 unit rather than split by layer: unlike Hohle Fels, its documented mixing is *between* the Aurignacian and other technocomplexes (Gravettian, Magdalenian), not between Aur-P1 and Aur-P2 Aurignacian facies, so no layer-level phase boundary can be drawn from the published reassessments. The mixing concern is instead addressed directly in Supplement S8 Analysis B (exclusion of Vogelherd).

- @Schurch_Conard_2026_reassess : Reassesses cultural stratigraphy; AH IV–V are Aurignacian but with severe mixing (38 refit connections between layers); 15 radiocarbon dates span 42–34.5 ka BP (calibrated); Gravettian also present.
- @Schurch_et_al_2025_bladelet : Blade/bladelet production analysis; mixing documented but majority of artifacts are Aurignacian.
- Original dating of Vogelherd Aurignacian to 41–35 ka BP (calibrated).

### Fumane → Protoaurignacian

- @Falcucci_Peresani_2017 : Protoaurignacian techno-typology persists throughout sequence at Fumane (A2-A1 through D3).
- @falcucci2020breaking : "Features of the Protoaurignacian techno-typology are present throughout the stratigraphic sequence."

### La Ferrassie → Early Aurignacian

- @Talamo_La_Ferrassie_2020 : Technological attribution places La Ferrassie in the Early Aurignacian.

### Abri Pataud → Early Aurignacian

- @Chiotti_2005 : Technological analysis assigns Abri Pataud to the Early Aurignacian.
- @Higham_Abri_Pataud_2011 : Radiocarbon dating supports Early Aurignacian placement.

### Mladeč → Evolved Aurignacian

- @Oliva_2006 : Technological and typological analysis assigns Mladeč to the Evolved Aurignacian.
- @Wild_et_al_2005 : Direct AMS dates of ~31 ka ¹⁴C BP (~35–36 ka cal BP) support an Evolved Aurignacian attribution.

### Spy → Excluded

- @Flas_et_al_2013 : The Spy Aurignacian assemblage exhibits features of both Early and Late Aurignacian phases. Split-based osseous points suggest Early Aurignacian; small curved bladelets from nosed endscrapers, burins busqués, and Vachons burins suggest Late Aurignacian.
- @Flas_2015 : The Aurignacian at Spy is a mixed assemblage containing a small Early Aurignacian component and several Late Aurignacian components. The earliest convincing Aurignacian date in the region is about 33,000–32,000 BP.
- @Semal_2009 : The Spy Neandertals date to ~36,000 BP, associated with the Lincombian-Ranisian-Jerzmanowician (LRJ). The LRJ and Aurignacian are mixed in the "deuxième niveau ossifère."
- The site is heavily mixed: Multiple excavations (1885–1886, 1903–1904, 1952–1960s) with inconsistent methods. The "deuxième niveau ossifère" contains Mousterian, LRJ, and Aurignacian materials mixed together.

### Gatzarria → Proto-Aurignacian

- @Barshay_Szmidt_2012 : Layer Cjn2 is a Proto-Aurignacian — "fits completely within the definition and characterization of the Proto-Aurignacian" with exclusive blade/bladelet production and Dufour bladelets (sub-type Dufour). Layer Cjn1 may be mixed with Cjn2 and Cbf (Classic Aurignacian). The back of the cave is mixed, but the front zone ("sampled set") is internally coherent; refitting confirms minimal vertical displacement.
- @Deschamps_2019 : Confirms complex stratigraphy and mixing concerns, but notes Cjn2 is internally coherent in the anterior zone.

### El Castillo → Excluded

- Mixed/contaminated layers preclude reliable phase assignment.

## Site status table

`signBase_Version1.0.csv` contains **`r nrow(signbase_full)` objects distributed across `r n_distinct(signbase_full$site_name)` sites**. `r n_zero_objects` objects from `r n_zero_sites` sites carry no geometric signs and are excluded upstream. A further eight sites are removed from the sample before phase assignment (Willendorf; Muralovka, Shanidar Cave and Hayonim Cave for geographic isolation; El Salitre for lack of a reliable chronological context; Riparo di Fontana Nuova and Grotte De La Princesse Pauline, and Šandalja II as later or peripheral). After phase reclassification and the exclusion of the two mixed-assemblage sites Spy and El Castillo, the analysed sample retains **`r n_retained_sites` sites carrying `r n_retained_objects` objects across `r n_retained_countries` countries**: Aur-P1 (Proto/Early) has `r n_p1_sites` sites and `r n_p1_objects` objects, Aur-P2 (Evolved/Late) has `r n_p2_sites` sites and `r n_p2_objects` objects.

```{r}
#| label: site-table
#| echo: false
site_table <- signbase_full_clean %>%
  group_by(site_name, phase2) %>%
  summarise(objects = n(), .groups = "drop") %>%
  arrange(site_name)

knitr::kable(site_table, col.names = c("Site", "Phase", "Objects"),
             caption = "Sites and object counts after phase reclassification (Aur-P1 = Proto/Early, Aur-P2 = Evolved/Late).")
```

### Aur-P1 (Proto/Early) — technology-based assignment

Sites placed in Aur-P1 on the basis of specialist technological attributions rather than radiocarbon dates alone.

```{r}
#| label: tbl-tech-aurp1
#| echo: false
tech_p1 <- site_objects %>%
  filter(assignment == "technology", phase2 == "Aur-P1") %>%
  arrange(desc(objects)) %>%
  select(site_name, objects, basis)

knitr::kable(tech_p1, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P1 sites assigned by technology rather than radiocarbon date.")
```

### Aur-P2 (Evolved/Late) — technology-based assignment

```{r}
#| label: tbl-tech-aurp2
#| echo: false
tech_p2 <- site_objects %>%
  filter(assignment == "technology", phase2 == "Aur-P2") %>%
  select(site_name, objects, basis)

knitr::kable(tech_p2, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P2 sites assigned by technology rather than radiocarbon date.")
```

### Excluded due to mixing

```{r}
#| label: tbl-excluded
#| echo: false
excluded_table <- signbase_full %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  filter(site_name %in% excluded_basis$site_name, sign_total > 0) %>%
  group_by(site_name) %>%
  summarise(objects = n(), .groups = "drop") %>%
  left_join(excluded_basis, by = "site_name") %>%
  select(site_name, objects, basis)

knitr::kable(excluded_table, col.names = c("Site", "Objects", "Basis"),
             caption = "Sites excluded from the analysis because of mixed/contaminated assemblages.")
```

### Aur-P1 (Proto/Early) — date-based assignment

Sites assigned by the calibrated date thresholds in the `assign-phases` chunk (default case), which fall in Aur-P1 once the Proto/Early distinction is collapsed. Sites listed here are unchanged from a date-only assignment.

```{r}
#| label: tbl-date-aurp1
#| echo: false
date_p1 <- site_objects %>%
  filter(assignment == "date", phase2 == "Aur-P1") %>%
  arrange(desc(objects)) %>%
  select(site_name, objects, basis)

knitr::kable(date_p1, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P1 sites assigned by their calibrated radiocarbon date.")
```

### Aur-P2 (Evolved/Late) — date-based assignment

```{r}
#| label: tbl-date-aurp2
#| echo: false
date_p2 <- site_objects %>%
  filter(assignment == "date", phase2 == "Aur-P2") %>%
  arrange(desc(objects)) %>%
  select(site_name, objects, basis)

knitr::kable(date_p2, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P2 sites assigned by their calibrated radiocarbon date.")
```

### Objects without geometric signs

These are **individual objects**, not sites: 15 objects from 13 sites carry no geometric signs and are removed upstream by the `sign_total > 0` filter. A site's signed objects remain in the analysis only if the site also passes the date-parsing/calibration step (27 sites drop out there) and phase assignment.

```{r}
#| label: tbl-zero-sign
#| echo: false
zero_sign_summary <- signbase_full %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  group_by(site_name) %>%
  summarise(
    objects_without_signs = sum(sign_total == 0),
    signed_objects = sum(sign_total > 0),
    .groups = "drop"
  ) %>%
  filter(objects_without_signs > 0) %>%
  arrange(desc(objects_without_signs)) %>%
  mutate(
    in_analysis = ifelse(site_name %in% signbase_full_clean$site_name,
                         "Yes",
                         ifelse(site_name %in% c("Spy", "El Castillo"),
                                "No (excluded for mixing)",
                                ifelse(site_name %in% c("Willendorf", "Riparo di Fontana Nuova", "Muralovka",
                                                       "Shanidar Cave", "Hayonim Cave", "El Salitre",
                                                       "Grotte De La Princesse Pauline", "Šandalja II"),
                                       "No (excluded before phase assignment)",
                                       "No (date parsing)")))
  )

knitr::kable(zero_sign_summary,
             col.names = c("Site", "Objects without signs", "Signed objects", "In analysis?"),
             caption = "Sites contributing objects without geometric signs, and whether the site remains in the analysis.")
```

Only **Willendorf** and **Spy** drop out entirely; the remaining "No" sites are lost at the date-parsing/calibration step rather than because of the zero-sign filter.

## References cited


