---
title: "Supplementary Material: Site Phase Assignments"
subtitle: "Geometric signs as proxies for cultural connectivity during the Western Eurasian Aurignacian"
author:
  - Sophie Cooper:
      institute: uw
      orcid: xxx
  - Aleah Rosner:
      institute: uw
      orcid: xxx
  - Eva Oryn:
      institute: uw
      orcid: xxx
  - Kenneth B. Vernon:
      orcid: 0000-0003-0098-5092
      institute:
        - uu
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0001-7879-4531
      institute:
        - uw
institute:
  - uw: Department of Anthropology, University of Washington, Seattle, USA
  - uu: Scientific Computing and Computing Institute, University of Utah, Salt Lake City, USA
title-block-published: "Last updated"
date: now
date-format: long
format:
  docx:
    reference-doc: "../templates/template.docx"
execute:
  echo: true
  warning: false
  message: false
---

## Overview

This supplement documents the rationale and procedure for assigning archaeological sites to Aurignacian phases. Phase assignments in the main analysis are based on specialist technological attributions from the published literature, not solely on radiocarbon dates. This supplement summarises the evidence for each reclassification and provides the complete site-level status table.

## Data pipeline

The following code loads the raw SignBase dataset, parses radiocarbon dates, calibrates them, and assigns objects to phases. Sites excluded from the main analysis are filtered out, and sites reassigned based on technological attribution are overrides in the `case_when()`.

```{r}
#| label: load-data
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(rcarbon)
library(here)

signbase_full <- read.csv(here("data/signBase_Version1.0.csv"), header = TRUE)

signbase_full_clean <- signbase_full %>%
  filter(site_name != "Willendorf",
         site_name != "Riparo di Fontana Nuova",
         site_name != "Muralovka",
         site_name != "Shanidar Cave",
         site_name != "Hayonim Cave",
         site_name != "El Salitre",
         site_name != "Grotte De La Princesse Pauline",
         site_name != "Šandalja II") %>%
  dplyr::select(-other, -rectangle) %>%
  mutate(longitude = ifelse(site_name == "Riparo Bombrini", 7.437500, longitude)) %>%
  mutate(latitude = ifelse(site_name == "Riparo Bombrini", 43.77083, latitude))
```

```{r}
#| label: parse-dates
signbase_years <- signbase_full_clean %>%
  drop_na(date_bp_max_min) %>%
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+\\/\\-", "±")) %>%
  mutate(date_bp_max_min = str_replace_all(date_bp_max_min, "\\+", "±")) %>%
  separate(date_bp_max_min,
           sep = " - ",
           c("date_bp_max", "date_bp_min"),
           remove = FALSE) %>%
  mutate(date_bp_max = ifelse(str_detect(date_bp_max, "\\/"),
                              str_extract(date_bp_max, ".*?(?=\\/)"),
                              date_bp_max)) %>%
  mutate(date_bp_min = ifelse(str_detect(date_bp_min, "\\/"),
                              str_extract(date_bp_min, ".*?(?=\\/)"),
                              date_bp_min)) %>%
  separate(date_bp_max,
           sep = "±",
           c("date_bp_max_age",
             "date_bp_max_error")) %>%
  separate(date_bp_min,
           sep = "±",
           c("date_bp_min_age",
             "date_bp_min_error")) %>%
  drop_na(date_bp_max_age,
          date_bp_max_error) %>%
  mutate(date_bp_max_age = parse_number(date_bp_max_age),
         date_bp_max_error = parse_number(date_bp_max_error))

signbase_years_cal <-
  rcarbon::calibrate(signbase_years$date_bp_max_age,
                     signbase_years$date_bp_max_error,
                     verbose = FALSE) %>%
  summary() %>%
  tibble()

signbase_years$MedianBP <- signbase_years_cal$MedianBP

signbase_full_clean <- signbase_full_clean %>%
  inner_join(signbase_years %>%
              dplyr::select(object_id, MedianBP))
```

```{r}
#| label: assign-phases
# Date-based phase thresholds (conventional round numbers)
# Proto-Aurignacian:  > 39,799 BP
# Early Aurignacian:  37,799–39,799 BP
# Evolved Aurignacian: 31,999–37,799 BP
# These three-phase labels are then collapsed to the two-phase (Aur-P1/Aur-P2)
# scheme used in the main analysis: Aur-P1 = Proto + Early, Aur-P2 = Evolved/Late.

signbase_full_clean <- signbase_full_clean %>%
  mutate(time_period = case_when(
    # --- Exclusions (mixed assemblages) ---
    site_name == "Spy"         ~ NA_character_,
    site_name == "El Castillo" ~ NA_character_,
    # --- Reclassifications based on technological attribution ---
    # Hohle Fels is assigned at the layer level: the lower Aurignacian horizons
    # (AH Vb, Vab, Vaa, Va) are Early Aurignacian (Aur-P1), while the upper
    # horizons (AH IV, IIIb, IIIa and the II-series) are Evolved/Late
    # Aurignacian (Aur-P2), following Dinnis et al. (2019) and Bataille &
    # Conard (2018). The single object from the mixed layer IIIa-V spans the
    # whole sequence and is excluded.
    site_name == "Hohle Fels" & layer %in% c("Va", "Vaa", "Vab", "Vb") ~ "early_aurignacian",
    site_name == "Hohle Fels" & layer %in% c("IV", "IIIa", "IIIb", "Iid", "Iida", "Iie") ~ "evolved_aurignacian",
    site_name == "Hohle Fels" & layer == "IIIa-V" ~ NA_character_,
    site_name == "Geissenklösterle"  ~ "early_aurignacian",
    site_name == "Vogelherd"         ~ "early_aurignacian",
    site_name == "La Ferrassie"      ~ "early_aurignacian",
    site_name == "Abri Pataud"       ~ "early_aurignacian",
    site_name == "Mladeč"            ~ "evolved_aurignacian",
    # --- Override: Gatzarria Cjn2 is Proto-Aurignacian ---
    site_name == "Gatzarria"         ~ "proto_aurignacian",
    # --- Default: date-based assignment ---
    MedianBP > 39799 ~ "proto_aurignacian",
    MedianBP > 37799 ~ "early_aurignacian",
    MedianBP > 31999 ~ "evolved_aurignacian"
  )) %>%
  filter(!is.na(time_period)) %>%
  mutate(phase2 = ifelse(time_period %in% c("proto_aurignacian", "early_aurignacian"),
                         "Aur-P1", "Aur-P2")) %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  filter(sign_total > 0)
```

```{r}
#| label: summary-counts
site_summary <- signbase_full_clean %>%
  group_by(phase2) %>%
  summarise(
    sites = n_distinct(site_name),
    objects = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(objects))

site_summary
```

```{r}
#| label: site-basis
# Literature-rationale text ONLY: the site names and object counts come from
# the pipeline above; only the "Basis" citation strings are recorded by hand.
# Any site not listed here is date-assigned (the default case_when branch).
site_basis <- data.frame(
  site_name = c("Hohle Fels", "Geissenklösterle", "Vogelherd", "La Ferrassie",
                "Abri Pataud", "Gatzarria", "Mladeč"),
  assignment = "technology",
  basis = c(
    "Layer-level split: AH V Early Aurignacian (Aur-P1); AH IV, III and II-series Evolved/Late Aurignacian (Aur-P2) (Bataille & Conard 2018, 2019; Dinnis et al. 2019)",
    "Early Aurignacian (Bataille & Conard 2018)",
    "Early Aurignacian (Schürch & Conard 2026; Bataille & Conard 2018)",
    "Early Aurignacian (Talamo et al. 2020)",
    "Early Aurignacian (Chiotti 2005; Higham et al. 2011)",
    "Proto-Aurignacian, layer Cjn2 (Barshay-Szmidt et al. 2012)",
    "Evolved Aurignacian (Oliva 2006)"
  )
)

excluded_basis <- data.frame(
  site_name = c("Spy", "El Castillo"),
  assignment = "excluded",
  basis = c(
    "Mixed assemblage, no single Proto-Aurignacian attribution (Flas et al. 2013; Flas 2015; Semal et al. 2009)",
    "Mixed/contaminated layers preclude reliable assignment"
  )
)

# Object counts per retained site, tagged by assignment basis
site_objects <- signbase_full_clean %>%
  group_by(site_name, phase2) %>%
  summarise(objects = n(), median_bp = median(MedianBP), .groups = "drop") %>%
  left_join(site_basis, by = "site_name") %>%
  mutate(assignment = ifelse(is.na(assignment), "date", assignment))

# Date-assigned sites get the threshold string generated from their own MedianBP
site_objects <- site_objects %>%
  mutate(basis = ifelse(
    assignment == "date",
    case_when(
      median_bp > 39799 ~ "MedianBP > 39,799",
      median_bp > 37799 ~ "MedianBP 37,799\u201339,799",
      TRUE               ~ "MedianBP 31,999\u201337,799"
    ),
    basis
  )) %>%
  mutate(basis = ifelse(
    site_name == "Fumane",
    "MedianBP > 39,799 (Protoaurignacian throughout sequence; Falcucci et al. 2017, 2020)",
    basis
  ))

knitr::kable(site_objects, caption = "Retained sites with object counts and assignment basis.")
```

```{r}
#| label: numeric-summary
# Inline-prose numbers computed from the pipeline (never hard-coded).
zero_sign_objects <- signbase_full %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  filter(sign_total == 0)

n_zero_objects <- nrow(zero_sign_objects)
n_zero_sites  <- n_distinct(zero_sign_objects$site_name)

n_retained_sites    <- n_distinct(signbase_full_clean$site_name)
n_retained_objects  <- nrow(signbase_full_clean)
n_retained_countries <- n_distinct(signbase_full_clean$country)

n_p1_sites   <- site_summary[["sites"]][site_summary$phase2 == "Aur-P1"]
n_p1_objects <- site_summary[["objects"]][site_summary$phase2 == "Aur-P1"]
n_p2_sites   <- site_summary[["sites"]][site_summary$phase2 == "Aur-P2"]
n_p2_objects <- site_summary[["objects"]][site_summary$phase2 == "Aur-P2"]
```

## Key literature findings

### Hohle Fels → layer-level assignment (AH V Early; AH IV, III, II-series Evolved)

Hohle Fels is the only site in the dataset whose internal sequence warrants splitting the phase assignment at the layer level rather than treating the whole site as a single phase.

- Bataille & Conard 2018 (PLoS ONE): AH IV is a "techno-functional facies of the Swabian Aurignacian" with burin-core bladelet production; it shares features with the Early Aurignacian but is compositionally distinct from the underlying AH V assemblage, challenging the straightforward application of the Western European typo-chronological system to Central Europe.
- Bataille & Conard 2019 (Quartär): AH IIIa and IV share similar tool/core assemblages typical of the Swabian Aurignacian, forming a coherent upper unit that contrasts with AH V.
- Dinnis et al. 2019 (PPS): Hohle Fels IV shows "important similarities with the Western European Late Aurignacian"; its radiocarbon dates of 33–30 ka uncal BP are not older than those of the Western European Late Aurignacian. This is the direct basis for placing AH IV and the overlying III and II-series layers in Aur-P2 (Evolved/Late).
- AH V (Va, Vaa, Vab, Vb) is the earlier Aurignacian unit at Hohle Fels and is retained in Aur-P1 (Early Aurignacian); its calibrated median dates (~39.7–43.3 ka BP) are consistent with that placement.
- The single object recorded from the mixed layer "IIIa-V" spans the whole Aurignacian sequence and is therefore excluded from the analysis, in line with the treatment of the mixed-assemblage sites Spy and El Castillo.

### Geissenklösterle → Early Aurignacian (Aur-P1)

- Bataille & Conard 2018 (PLoS ONE): The Geissenklösterle Aurignacian (AH III and II) is assigned to the Early Aurignacian on technological grounds.
- Both Geissenklösterle layers remain in Aur-P1: the calibrated median dates of the SignBase radiocarbon determinations for AH II (~40.7 ka BP) and AH III (~42.7 ka BP) fall within the Aur-P1 range, so the technology- and date-based assignments agree for this site. A minority reassessment (Zilhão 2013) places AH III later (~33–35 ka BP), but this conflicts with the site's own radiocarbon chronology and is not adopted here; the weight of the technological and stratigraphic evidence keeps both layers in Aur-P1.

### Vogelherd → Early Aurignacian (Aur-P1, site-level)

Vogelherd is kept as a single site-level Aur-P1 unit rather than split by layer: unlike Hohle Fels, its documented mixing is *between* the Aurignacian and other technocomplexes (Gravettian, Magdalenian), not between Aur-P1 and Aur-P2 Aurignacian facies, so no layer-level phase boundary can be drawn from the published reassessments. The mixing concern is instead addressed directly in Supplement S8 Analysis B (exclusion of Vogelherd).

- Schürch & Conard 2026 (JPA): Reassesses cultural stratigraphy; AH IV–V are Aurignacian but with severe mixing (38 refit connections between layers); 15 radiocarbon dates span 42–34.5 ka BP (calibrated); Gravettian also present.
- Schürch et al. 2025 (PLoS ONE): Blade/bladelet production analysis; mixing documented but majority of artifacts are Aurignacian.
- Conard & Bolus 2003, 2008: Original dating of Vogelherd Aurignacian to 41–35 ka BP (calibrated).

### Fumane → Protoaurignacian

- Falcucci et al. 2017 (PLoS ONE): Protoaurignacian techno-typology persists throughout sequence at Fumane (A2-A1 through D3).
- Falcucci et al. 2020 (JASS): "Features of the Protoaurignacian techno-typology are present throughout the stratigraphic sequence."
- Falcucci 2018 (MGFu): "Protoaurignacian clearly persists throughout the stratigraphic sequence."

### La Ferrassie → Early Aurignacian

- Talamo et al. 2020: Technological attribution places La Ferrassie in the Early Aurignacian.

### Abri Pataud → Early Aurignacian

- Chiotti 2005: Technological analysis assigns Abri Pataud to the Early Aurignacian.
- Higham et al. 2011: Radiocarbon dating supports Early Aurignacian placement.

### Mladeč → Evolved Aurignacian

- Oliva 2006: Technological and typological analysis assigns Mladeč to the Evolved Aurignacian.

### Spy → Excluded

- Flas et al. 2013: The Spy Aurignacian assemblage exhibits features of both Early and Late Aurignacian phases. Split-based osseous points suggest Early Aurignacian; small curved bladelets from nosed endscrapers, burins busqués, and Vachons burins suggest Late Aurignacian.
- Flas 2015: The Aurignacian at Spy is a mixed assemblage containing a small Early Aurignacian component and several Late Aurignacian components. The earliest convincing Aurignacian date in the region is about 33,000–32,000 BP.
- Semal et al. 2009: The Spy Neandertals date to ~36,000 BP, associated with the Lincombian-Ranisian-Jerzmanowician (LRJ). The LRJ and Aurignacian are mixed in the "deuxième niveau ossifère."
- The site is heavily mixed: Multiple excavations (1885–1886, 1903–1904, 1952–1960s) with inconsistent methods. The "deuxième niveau ossifère" contains Mousterian, LRJ, and Aurignacian materials mixed together.

### Gatzarria → Proto-Aurignacian

- Barshay-Szmidt et al. 2012: Layer Cjn2 is a Proto-Aurignacian — "fits completely within the definition and characterization of the Proto-Aurignacian" with exclusive blade/bladelet production and Dufour bladelets (sub-type Dufour). Layer Cjn1 may be mixed with Cjn2 and Cbf (Classic Aurignacian). The back of the cave is mixed, but the front zone ("sampled set") is internally coherent; refitting confirms minimal vertical displacement.
- Deschamps 2019: Confirms complex stratigraphy and mixing concerns, but notes Cjn2 is internally coherent in the anterior zone.

### El Castillo → Excluded

- Mixed/contaminated layers preclude reliable phase assignment.

## Site status table

`signBase_Version1.0.csv` contains **`r nrow(signbase_full)` objects distributed across `r n_distinct(signbase_full$site_name)` sites**. `r n_zero_objects` objects from `r n_zero_sites` sites carry no geometric signs and are excluded upstream. A further eight sites are removed from the sample before phase assignment (Willendorf; Muralovka, Shanidar Cave and Hayonim Cave for geographic isolation; El Salitre for lack of a reliable chronological context; Riparo di Fontana Nuova and Grotte De La Princesse Pauline, and Šandalja II as later or peripheral). After phase reclassification and the exclusion of the two mixed-assemblage sites Spy and El Castillo, the analysed sample retains **`r n_retained_sites` sites carrying `r n_retained_objects` objects across `r n_retained_countries` countries**: Aur-P1 (Proto/Early) has `r n_p1_sites` sites and `r n_p1_objects` objects, Aur-P2 (Evolved/Late) has `r n_p2_sites` sites and `r n_p2_objects` objects.

```{r}
#| label: site-table
#| echo: false
site_table <- signbase_full_clean %>%
  group_by(site_name, phase2) %>%
  summarise(objects = n(), .groups = "drop") %>%
  arrange(site_name)

knitr::kable(site_table, col.names = c("Site", "Phase", "Objects"),
             caption = "Sites and object counts after phase reclassification (Aur-P1 = Proto/Early, Aur-P2 = Evolved/Late).")
```

### Aur-P1 (Proto/Early) — technology-based assignment

Sites placed in Aur-P1 on the basis of specialist technological attributions rather than radiocarbon dates alone.

```{r}
#| label: tbl-tech-aurp1
#| echo: false
tech_p1 <- site_objects %>%
  filter(assignment == "technology", phase2 == "Aur-P1") %>%
  arrange(desc(objects)) %>%
  select(site_name, objects, basis)

knitr::kable(tech_p1, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P1 sites assigned by technology rather than radiocarbon date.")
```

### Aur-P2 (Evolved/Late) — technology-based assignment

```{r}
#| label: tbl-tech-aurp2
#| echo: false
tech_p2 <- site_objects %>%
  filter(assignment == "technology", phase2 == "Aur-P2") %>%
  select(site_name, objects, basis)

knitr::kable(tech_p2, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P2 sites assigned by technology rather than radiocarbon date.")
```

### Excluded due to mixing

```{r}
#| label: tbl-excluded
#| echo: false
excluded_table <- signbase_full %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  filter(site_name %in% excluded_basis$site_name, sign_total > 0) %>%
  group_by(site_name) %>%
  summarise(objects = n(), .groups = "drop") %>%
  left_join(excluded_basis, by = "site_name") %>%
  select(site_name, objects, basis)

knitr::kable(excluded_table, col.names = c("Site", "Objects", "Basis"),
             caption = "Sites excluded from the analysis because of mixed/contaminated assemblages.")
```

### Aur-P1 (Proto/Early) — date-based assignment

Sites assigned by the calibrated date thresholds in the `assign-phases` chunk (default case), which fall in Aur-P1 once the Proto/Early distinction is collapsed. Sites listed here are unchanged from a date-only assignment.

```{r}
#| label: tbl-date-aurp1
#| echo: false
date_p1 <- site_objects %>%
  filter(assignment == "date", phase2 == "Aur-P1") %>%
  arrange(desc(objects)) %>%
  select(site_name, objects, basis)

knitr::kable(date_p1, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P1 sites assigned by their calibrated radiocarbon date.")
```

### Aur-P2 (Evolved/Late) — date-based assignment

```{r}
#| label: tbl-date-aurp2
#| echo: false
date_p2 <- site_objects %>%
  filter(assignment == "date", phase2 == "Aur-P2") %>%
  arrange(desc(objects)) %>%
  select(site_name, objects, basis)

knitr::kable(date_p2, col.names = c("Site", "Objects", "Basis"),
             caption = "Aur-P2 sites assigned by their calibrated radiocarbon date.")
```

### Objects without geometric signs

These are **individual objects**, not sites: 15 objects from 13 sites carry no geometric signs and are removed upstream by the `sign_total > 0` filter. A site's signed objects remain in the analysis only if the site also passes the date-parsing/calibration step (27 sites drop out there) and phase assignment.

```{r}
#| label: tbl-zero-sign
#| echo: false
zero_sign_summary <- signbase_full %>%
  mutate(sign_total = rowSums(.[, 25:53])) %>%
  group_by(site_name) %>%
  summarise(
    objects_without_signs = sum(sign_total == 0),
    signed_objects = sum(sign_total > 0),
    .groups = "drop"
  ) %>%
  filter(objects_without_signs > 0) %>%
  arrange(desc(objects_without_signs)) %>%
  mutate(
    in_analysis = ifelse(site_name %in% signbase_full_clean$site_name,
                         "Yes",
                         ifelse(site_name %in% c("Spy", "El Castillo"),
                                "No (excluded for mixing)",
                                ifelse(site_name %in% c("Willendorf", "Riparo di Fontana Nuova", "Muralovka",
                                                       "Shanidar Cave", "Hayonim Cave", "El Salitre",
                                                       "Grotte De La Princesse Pauline", "Šandalja II"),
                                       "No (excluded before phase assignment)",
                                       "No (date parsing)")))
  )

knitr::kable(zero_sign_summary,
             col.names = c("Site", "Objects without signs", "Signed objects", "In analysis?"),
             caption = "Sites contributing objects without geometric signs, and whether the site remains in the analysis.")
```

Only **Willendorf** and **Spy** drop out entirely; the remaining "No" sites are lost at the date-parsing/calibration step rather than because of the zero-sign filter.
